---
title: "SBTD_Tissue"
author: "Elise Stacy"
date: "2025-08-26"
output: html_document
---
# Load packages
```{r}
# Additional filtering
library(vcfR)
library(SNPfiltR) 

library(adegenet) # Locus div calcs
library(pegas) # For HWE calcs

# Data structure manipulation
library(stringr)
library(purrr)
library(dplyr)
library(tidyr)

# Plotting
library(ggplot2)
library(gridExtra)
library(maps)
```
## Load error calculation functions (code slightly modified from https://github.com/arianacerreta)
```{r}
## Scoring matching genotypes
#make a function
score_gt_match<-function(gt1,gt2){
  #Handle missing genotypes
  if (is.na(gt1) & is.na(gt2)) return("BothNA")
  else if (is.na(gt1) | is.na(gt2)) return("OneNA")
  
  #split alleles
  a1 <- unlist(strsplit(gt1, "[/|]"))
  a2 <- unlist(strsplit(gt2, "[/|]"))
  
  #sort alleles for unphased comparison
  a1<- sort(a1)
  a2<- sort(a2)
  
  #score
  if (identical(a1,a2)) return("Full Match")
  else if (any(a1 %in% a2)) return("Partial Match")
  else return("Mismatch")
}

format_paired_gt_df <- function(vcfR_samples, popmap_samples) {

  # Extract base replicate IDs
  popmap_samples <- popmap_samples %>%
    mutate(ReplicateID = sub(".*_", "", id))

  # Keep only replicate groups with >=2 samples
  replicate_counts <- popmap_samples %>%
    count(ReplicateID) %>%
    filter(n >= 2)

  # Filter popmap_samples to those with replicates
  popmap_samples <- popmap_samples %>%
    filter(ReplicateID %in% replicate_counts$ReplicateID)

  # Subset VCF to matching sample names
  # Ensure only selected samples are kept
  sample_ids <- popmap_samples$id
  vcfR_samples@gt <- vcfR_samples@gt[, c("FORMAT", sample_ids)]


  # Extract GT matrix
  gt_df <- extract.gt(vcfR_samples, element = "GT", as.numeric = FALSE) %>% as.data.frame()
  gt_df$CHROM <- getCHROM(vcfR_samples)
  gt_df$POS <- as.numeric(getPOS(vcfR_samples))
  gt_df <- gt_df %>% relocate(CHROM, POS)

  # Pivot to long format
  gt_df <- pivot_longer(gt_df, cols = -c(CHROM, POS), names_to = "SampleID", values_to = "GT")
  gt_df <- gt_df %>%
    mutate(ReplicateID = sub(".*_", "", SampleID))

  # Generate all pairwise combinations of replicates
  replicate_map <- popmap_samples %>%
    select(ReplicateID, id) %>%
    group_by(ReplicateID) %>%
    filter(n() >= 2) %>%
    summarise(pair_df = list(as.data.frame(t(combn(id, 2)))), .groups = "drop") %>%
    unnest(pair_df) %>%
    rename(Sample1 = V1, Sample2 = V2)


  # Join with genotypes
  paired_gt_df <- replicate_map %>%
    left_join(gt_df, by = c("Sample1" = "SampleID")) %>%
    rename(GT1 = GT, CHROM = CHROM, POS = POS) %>%
    left_join(gt_df, by = c("Sample2" = "SampleID", "CHROM", "POS")) %>%
    rename(GT2 = GT)

  return(paired_gt_df)
}

# Process data and create summaries in one function
calculate_gt_pairs_errors <- function(paired_gt_df) {
  # Create gt_pairs_mismatch dataframe
  gt_pairs_mismatch <- paired_gt_df %>%
    rowwise() %>%
    mutate(MatchType = score_gt_match(GT1, GT2)) %>%
    ungroup()
  
  # Calculate summary_by_pair
  summary_by_pair <- gt_pairs_mismatch %>%
    group_by(Sample1, Sample2) %>%
    summarize(
      N = n(),
      Full_Match = sum(MatchType == "Full Match", na.rm = TRUE),
      Partial_Match = sum(MatchType == "Partial Match", na.rm = TRUE),
      Mismatch = sum(MatchType == "Mismatch", na.rm = TRUE),
      Missing_Full = sum(MatchType == "BothNA", na.rm = TRUE),
      Missing_Partial = sum(MatchType == "OneNA", na.rm = TRUE),
      Full_Match_Rate = Full_Match / (Full_Match + Partial_Match + Mismatch),
      Mismatch_Rate = (Mismatch + Partial_Match) / (Full_Match + Partial_Match + Mismatch),
      Full_Missingness = Missing_Full / N,
      Partial_Missingness = Missing_Partial / N,
      Inclusive_Missingness = (Missing_Full + Missing_Partial) / N
    )
  
  # Calculate summary_by_locus
  summary_by_locus <- gt_pairs_mismatch %>%
    group_by(CHROM, POS) %>%
    summarize(
      N=n(),
      Full_Match = sum(MatchType == "Full Match", na.rm = TRUE),
      Mismatch = sum(MatchType == "Mismatch", na.rm = TRUE),
      Partial_Match = sum(MatchType == "Partial Match", na.rm = TRUE),
      Missing_Full = sum(MatchType == "BothNA", na.rm = TRUE),
      Missing_Partial = sum(MatchType == "OneNA", na.rm = TRUE),
      Full_Match_Rate = Full_Match / (Full_Match + Partial_Match + Mismatch),
      Both_Mismatch_Rate = (Partial_Match + Mismatch) / (Full_Match + Partial_Match + Mismatch),
      Full_Missingness = Missing_Full / N,
      Partial_Missingness = Missing_Partial / N,
      Inclusive_Missingness = (Missing_Full + Missing_Partial) / N
    )
  
  # Return both summaries
  list(summary_by_pair = summary_by_pair,
       summary_by_locus = summary_by_locus)
}
```
# Read in sample data
```{r}
snps_vcfR <- read.vcfR("/mnt/ceph/estacy/SBTDeer/Tissue_samples/05_filter/populations_R0.5_mac2/populations.snps.vcf")

locations <- read.csv("/mnt/ceph/estacy/SBTDeer/R_analysis/data/Tissue_locations.csv", na.strings = c("","NA"))

locations$Region[locations$Region == "Graham Island (Haida Gwaii)"] <- "Graham Island"

# Just present broad regions - split regions to northern Southeast, mid Southeast, southern Southeast

library(dplyr)

locations <- locations %>%
  mutate(
    Populations = case_when(
      Region %in% c("Gravina Island",
                    "Revillagigedo Island",
                    "Cleveland Peninsula") ~ "southern Southeast",
      Region %in% c("Kupreanof Island", "Mitkof Island") ~ "middle Southeast",
      Region == "Prince of Wales Island" ~ "P. of Wales Is.",
      Broad_region == "northern Southeast" ~ "northern Southeast",
      Broad_region == "Kodiak Archipelago" ~ "Kodiak",
      Broad_region == "Southcentral" ~ "Southcentral",
      Broad_region == "British Columbia" ~ "British Columbia",
      Broad_region == "Haida Gwaii" ~ "Haida Gwaii",
      Broad_region == NA ~ "Unknown"
    )
  )
```

# Subset (whitelist) to designed panel loci
```{r}
panel_snps <- read.csv("/mnt/ceph/estacy/SBTDeer/R_analysis/data/Eacker_Deer_248_Loci_Names.csv", header = F)

# Whitelist snps panel was designed for only
panel_snps <- panel_snps %>%
  separate(V1, into = c("CHROM", "POS"), sep = "_")

# Create combined keys
vcf_keys <- paste(snps_vcfR@fix[, "CHROM"], snps_vcfR@fix[, "POS"], sep = "_")
panel_keys <- paste(panel_snps$CHROM, panel_snps$POS, sep = "_")

# Filter based on matching keys
vcfR_panel <- snps_vcfR[vcf_keys %in% panel_keys, ]
```
# Filtering to compare schemes
## Error rates in replicates
```{r}
vcfR <- vcfR_panel

# Initialize storage data frames
pair_errors <- data.frame()
locus_errors <- data.frame()

# Loop through depth values
for (depth in seq(5,30, by = 5)) {
  # Loop through missingness from 80% through fully missing
    for (missing in seq(0.1,0.4, by = 0.05)) {
    
    # Apply hard filter with current depth and GQ
    vcfR_filtered <- hard_filter(vcfR = vcfR, depth = depth, gq = 30)
    vcfR_missing <- missing_by_sample(vcfR = vcfR_filtered, cutoff = missing)
    
    popmap_samples <- data.frame(id=colnames(vcfR_missing@gt)[2:length(colnames(vcfR_missing@gt))])

    # Format paired genotype data frame
    paired_gt_df <- format_paired_gt_df(vcfR_missing, popmap_samples)
    
    # Calculate error rate
    error_rate <- calculate_gt_pairs_errors(paired_gt_df)
    
    # Add current Depth and GQ values to error summaries
    pair_error <- cbind(error_rate$summary_by_pair, Depth = depth, Missing = missing)
    locus_error <- cbind(error_rate$summary_by_locus, Depth = depth, Missing = missing)
    
    # Accumulate error data
    pair_errors <- rbind(pair_errors, pair_error)
    locus_errors <- rbind(locus_errors, locus_error)
    }
}
```
# Visualize error rate by filtering scheme
```{r}
pair_errors$ID_1 <- sub(".*_", "", pair_errors$Sample1)
pair_errors$ID_2 <- sub(".*_", "", pair_errors$Sample2) 

pair_err_summ <- pair_errors %>% # Summarize mismatch data
  group_by(Depth, Missing) %>%
  summarise(
    N_Samples = n(),
    Avg_Mismatch = round(mean(Mismatch_Rate, na.rm = TRUE), 4),
    Max_Mismatch = round(max(Mismatch_Rate, na.rm = TRUE), 4),
    Min_Locus_Overlap = min(((1 - Inclusive_Missingness) * N), na.rm = TRUE),
    Avg_Missing = round(mean(Missing_Full, na.rm = TRUE), 1)
  )

png("figures/tissues_manuscript/Supp_error_rates.png", width = 6000, height = 6000, res = 600)
# Plot error rates
ggplot(pair_err_summ, aes(x=Missing, y = Depth, fill = Avg_Mismatch)) +
  geom_tile() +
  scale_fill_gradient2(low = "lightblue", mid ="purple", high = "red", midpoint = 0.0025) +
  geom_text(aes(label = N_Samples), color = "black", size = 3) +
  labs(title = "Sample replicate pairs retained and average mismatch", subtitle = "Across Missingness filtering") +
  xlab("Sample Missingness")
dev.off()
```
# Select one replicate per lab_id with the lowest missing_data and highest depth
```{r}
# Data frame of full unique IDs and lab IDs
sample_ids <- data.frame(id = sub(".*_", "", colnames(vcfR_panel@gt)[-1]), full_id = colnames(vcfR_panel@gt)[-1])
# Create set of replicate only ids
replicates <- sample_ids %>%
  group_by(id) %>%
  filter(n() > 1)
# Create vectors of replicate vs not rep samples
replicate_samples <- intersect(colnames(vcfR_panel@gt)[-1], replicates$full_id)
non_rep_samples <- setdiff(colnames(vcfR_panel@gt)[-1], replicates$full_id)

# Seperate vcfR objects in replicated samples and non rep objects
vcfR_reps <- vcfR_panel[samples = replicate_samples]
vcfR_non_reps <- vcfR_panel[samples = non_rep_samples]

# Apply locus filters
vcfR_reps_filt <- hard_filter(vcfR_reps, gq = 30, depth = 10)
vcfR_reps_filt <- min_mac(vcfR_reps_filt, min.mac = 2)

# Calculate missingness and depth by sample
missing_df_reps <- missing_by_sample(vcfR_reps_filt)
rep_stats <- missing_df_reps$unfiltered.stats
rep_stats$id <- sub(".*_", "", rep_stats$samples) # Format base sample ID only column

best_replicate <- rep_stats %>%
  group_by(id) %>%
  slice_min(prop.missing, with_ties = TRUE) %>% # Choose replicate with least missing data, allowing ties
  slice_max(mean.depth, with_ties = FALSE) %>% # Only paired should be ties now, choose replicate with greatest depth
  ungroup()

popmap_unique <- data.frame(id = c(best_replicate$samples, non_rep_samples))

vcfR_unique <- vcfR_panel[samples = popmap_unique$id]
```
# Panel locus filtering
```{r}
vcfR_dp <- hard_filter(vcfR_unique, gq = 30, depth = 10)
vcfR_missing <- missing_by_sample(vcfR_dp, cutoff = 0.2)
vcfR_mac <- min_mac(vcfR_missing, min.mac = 2)
vcfR_snp_miss <- missing_by_snp(vcfR_mac, cutoff = 0.5)

vcfR_panel_filt <- vcfR_snp_miss
```
### Mean depth
```{r}
dp <- extract.gt(vcfR_panel_filt, element = "DP", as.numeric=TRUE)

colMeans(dp, na.rm = T) %>% mean(na.rm = T)
colMeans(dp, na.rm = T) %>% min()
colMeans(dp, na.rm = T) %>% max()
```
# Format popmap
```{r}
popmap <- data.frame(id=colnames(vcfR_panel_filt@gt)[-1])

popmap_relabeled <- data.frame(id = sub(".*_", "", popmap$id))

sample_locations <- left_join(popmap_relabeled, locations, by = c("id"="UI.Barcode"))

popmap_relabeled$pop <- sample_locations$Populations

popmap_relabeled$pop <- factor(popmap_relabeled$pop, levels = c("Kodiak", "Southcentral", "northern Southeast", "middle Southeast", "southern Southeast", "P. of Wales Is.","British Columbia", "Haida Gwaii", "Unknown"))

colnames(vcfR_panel_filt@gt) <- c("FORMAT", popmap_relabeled$id)
```
## Map
### Format data to map
```{r}
locations$Populations <- factor(locations$Populations, levels = c("Kodiak", "Southcentral", "northern Southeast", "middle Southeast", "southern Southeast", "P. of Wales Is.", "British Columbia", "Haida Gwaii", "Unknown"))

# Sample size by location

locations_clean <- locations %>% filter(!is.na(Lon)) %>% filter(!is.na(UI.Barcode))

locations_clean <- locations_clean %>%
  group_by(Populations) %>%
  mutate(Population_n = paste(Populations," (n=", n(), ")", sep = "")) %>%
  ungroup()

SBTD_sites <- locations_clean[locations_clean$UI.Barcode %in% popmap_relabeled$id, ] # Plot samples with locations that passed filtering

labels_vector <- setNames(
  unique(SBTD_sites$Population_n),
  unique(SBTD_sites$Populations))

populations_color_pal <- c("#f9d5a9","#b5dee4","#3a7380","#eca255","#b45f06","#863e19","#682502","#431801","grey")
```
# Plot and save map of samples
```{r}
# get world map data
world_map_df <- map_data("world")

png("figures/tissues_manuscript/SBTD_sites_map.png", width = 6000, height = 6000, res = 600)

ggplot() +
  geom_polygon(data = world_map_df,
               aes(x = long, y = lat, group = group),
               fill = "grey98",
               colour = "grey50") +
  geom_point(data = locations_clean,
             aes(x = Lon, y = Lat,
                 fill = Populations),
             size = 4, shape = 21, stroke = .2) +
  scale_fill_manual(
    values = populations_color_pal,
    labels = labels_vector,
    guide = guide_legend(ncol = 1)
  ) +
  coord_quickmap(xlim = c(-155, -124), ylim = c(52.5, 62)) +
  theme_minimal() +
  labs(x = "Longitude", y = "Latitude") +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.text = element_text(size = 16),
    axis.text = element_text(size = 16),
    axis.title = element_text(size = 16)
  )

dev.off()
```
# Plot region popout map
```{r}
# Plot region map
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)

# Get USA states
states <- ne_states(country = "United States of America", returnclass = "sf")

# Get Canada provinces/territories (Admin-1)
canada <- ne_states(country = "Canada", returnclass = "sf")

# Combine for labeling
regions <- rbind(
  states %>% select(name, geometry),
  canada %>% select(name, geometry)
)

# Compute label positions
regions_centroids <- st_centroid(regions)

regions_centroids <- regions_centroids %>%
  mutate(
    label_x = st_coordinates(geometry)[,1],
    label_y = st_coordinates(geometry)[,2]
  )

# Move BC label slightly up and to the right, remove NWT
regions_centroids <- regions_centroids %>%
  mutate(
    label_x = ifelse(name == "British Columbia", label_x - 2.15, label_x),
    label_y = ifelse(name == "British Columbia", label_y + 3.75, label_y),
    label_name = ifelse(name == "British Columbia", "British\nColumbia", name)
  ) %>%
  filter(name != "Northwest Territories")  # <<< remove NWT label


regions_centroids <- regions_centroids %>%
  mutate(
    label_name = ifelse(name == "British Columbia", "British\nColumbia", name)
  )


png("figures/tissues_manuscript/region_map.png", width = 6000, height = 6000, res = 600)

ggplot() +
  geom_sf(data = ne_countries(scale = 50, returnclass = "sf"),
          fill = "grey", color = "grey70") +
  geom_sf(data = states, fill = NA, color = "#444444", size = 1.5) +
  geom_sf(data = canada, fill = NA, color = "#444444", size = 1.5) +
  geom_text(
    data = regions_centroids,
    aes(x = label_x, y = label_y, label = label_name),
    size = 14,
    fontface = "bold",
    color = "black"
  ) +
  coord_sf(xlim = c(-168, -119), ylim = c(49, 71.5), expand = FALSE) +
  theme_void() +  # <<< removes all background lines
  theme(
    text = element_text(size = 22)
  )

dev.off()
```
# PCA
```{r}
pca <- assess_missing_data_pca(vcfR=vcfR_panel_filt, popmap = popmap_relabeled, clustering = FALSE)
```
## Format for region labeled PCA figure
### Add sample size labels
```{r}
# Population order (controls legend order)
pop_order <- c(
  "Kodiak",
  "Southcentral",
  "northern Southeast",
  "middle Southeast",
  "southern Southeast",
  "P. of Wales Is.",
  "British Columbia",
  "Haida Gwaii",
  "Unknown"
)

pca$pop <- factor(pca$pop, levels = pop_order)

# Create population sample-size labels
pop_label_df <- pca %>%
  count(pop, name = "n") %>%
  mutate(label = paste0(pop, " (n=", n, ")"))

# Named vector of labels in pop_order
pop_labels_pca <- setNames(
  pop_label_df$label[match(pop_order, pop_label_df$pop)],
  pop_order
)

# Shape palette (named)
populations_shape_pal <- c(
  "Kodiak" = 21,
  "Southcentral" = 22,
  "northern Southeast" = 21,
  "middle Southeast" = 24,
  "southern Southeast" = 22,
  "P. of Wales Is." = 23,
  "British Columbia" = 25,
  "Haida Gwaii" = 21,
  "Unknown" = 23
)

# Color palette (named)
populations_color_pal_pca <- c(
  "Kodiak" = "#f9d5a9",
  "Southcentral" = "#b5dee4",
  "northern Southeast" = "#3a7380",
  "middle Southeast" = "#eca255",
  "southern Southeast" = "#b45f06",
  "P. of Wales Is." = "#863e19",
  "British Columbia" = "#682502",
  "Haida Gwaii" = "#431801",
  "Unknown" = "grey"
)
```
# Plot and save PCAs
```{r}
png("figures/tissues_manuscript/SBTD_pca.png", width = 6000, height = 6000, res = 600)

ggplot() +
  # All populations except BC & Unknown
  geom_point(
    data = subset(pca, !pop %in% c("British Columbia", "Unknown")),
    aes(x = PC1, y = PC2, fill = pop, shape = pop),
    size = 10, stroke = 0.5, alpha = 0.9
  ) +

  # BC & Unknown plotted on top
  geom_point(
    data = subset(pca, pop %in% c("British Columbia", "Unknown")),
    aes(x = PC1, y = PC2, fill = pop, shape = pop),
    size = 10, stroke = 0.5, alpha = 0.9
  ) +

  scale_fill_manual(
    values = populations_color_pal_pca,
    labels = pop_labels_pca,
    breaks = pop_order
  ) +

  scale_shape_manual(
    values = populations_shape_pal,
    labels = pop_labels_pca,
    breaks = pop_order
  ) +

  theme_light() +
  guides(
    fill  = guide_legend(order = 1, ncol = 2),
    shape = guide_legend(order = 1, ncol = 2)
  ) +

  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.text = element_text(size = 24),
    axis.text = element_text(size = 24),
    axis.title = element_text(size = 24)
  ) +

  labs(
    x = "PC1 12.91% variance explained",
    y = "PC2 5.29% variance explained"
  )

dev.off()

png("figures/tissues_manuscript/SBTD_pca_2_3.png", width = 6000, height = 6000, res = 600)

ggplot() +
  # 1. Plot all populations except Unknown
  geom_point(
    data = subset(pca, pop != "Unknown"),
    aes(x = PC2, y = PC3, fill = pop, shape = pop),
    size = 7, stroke = 0.5, alpha = 0.9
  ) +

  # 2. Plot Unknown samples on top
  geom_point(
    data = subset(pca, pop == "Unknown"),
    aes(x = PC1, y = PC2, fill = pop, shape = pop),
    size = 7, stroke = 0.5, alpha = 0.9
  ) +

  scale_fill_manual(
    values = populations_color_pal,
    labels = pop_labels_pca
  ) +

  # Add shape scaling
  scale_shape_manual(
    values = populations_shape_pal,    # <-- You must define this vector
    labels = pop_labels_pca
  ) +

  theme_light() +
  guides(
    fill = guide_legend(order = 1, ncol = 3),
    shape = guide_legend(order = 1, ncol = 3)
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.text = element_text(size = 16),
    axis.text = element_text(size = 14),
    axis.title = element_text(size = 16)
  #) +
  #labs(
  #  x = "PC1 XX% variance explained",
  #  y = "PC2 XX% variance explained"
  )

dev.off()
```
### Plot PCA by missingness
```{r}
png("figures/tissues_manuscript/SBTD_pca_missing_region.png", width = 6000, height = 3000, res = 600)

ggplot(data = pca, aes(x = PC1, y = PC2, color = pop, fill = missing)) +
  scale_color_manual(
    values = populations_color_pal,
    labels = pop_labels_pca) +
  scale_fill_gradient(
  low = "lightgrey",
  high = "black"
) +
  geom_point(size = 4, stroke = 1, shape = 21, alpha = 0.9) +
  theme_light() +
  theme()

dev.off()
```
# Relatedness 
Remove related individuals before Admixture
Running trioml on server, submission script: submit_related.sh and related.R
```{r}
library(related)

data(GenotypeData)
input <- readgenotypedata(GenotypeData)

ext.gt <- extract.gt(vcfR_panel_filt) %>% 
                                 t() %>%
                     as.data.frame() %>%
                     rownames_to_column(var = "ID")

gt_split <- ext.gt %>%
  mutate(across(-ID, ~strsplit(., "/"))) %>%
  unnest_wider(where(is.list), names_sep = "_") %>%
  mutate(across(-ID, as.numeric)) %>% 
  mutate(across(
    -ID,
    ~ case_when(
      is.na(.x) ~ 0,
      .x == 0   ~ 1,
      .x == 1   ~ 2,
      TRUE ~ .x    # keeps any unexpected values unchanged
    )
  ))

# Format one input file per population - running relatedness by population/region
gt_pop <- left_join(popmap_relabeled, gt_split, by = c("id" = "ID"))

gt_nSE <- gt_pop %>% filter(pop == "northern Southeast" | pop == "Unknown") %>% select(!pop)
gt_SC <- gt_pop %>% filter(pop == "Southcentral") %>% select(!pop)
gt_SE <- gt_pop %>% filter(pop == "middle Southeast" | pop == "southern Southeast" | pop == "British Columbia") %>% select(!pop)
gt_KA <- gt_pop %>% filter(pop == "Kodiak Archipelago") %>% select(!pop)
gt_HG <- gt_pop %>% filter(pop == "Haida Gwaii Archipelago") %>% select(!pop)

input_nSE <- readgenotypedata(gt_nSE)
input_SC <- readgenotypedata(gt_SC)
input_SE <- readgenotypedata(gt_SE)
input_KA <- readgenotypedata(gt_KA)
input_HG <- readgenotypedata(gt_HG)

saveRDS(input_nSE, "Relatedness/input_nSE.rds")
saveRDS(input_SC, "Relatedness/input_SC.rds")
saveRDS(input_SE, "Relatedness/input_SE.rds")
saveRDS(input_KA, "Relatedness/input_KA.rds")
saveRDS(input_HG, "Relatedness/input_HG.rds")
```
## Relatedness results and filter out first order relatedness
```{r}
HG_related <- readRDS(file = "Relatedness/output_HG_related.rds")
KA_related <- readRDS(file = "Relatedness/output_KA_related.rds")
nSE_related <- readRDS(file = "Relatedness/output_nSE_related.rds")
SE_related <- readRDS(file = "Relatedness/output_SE_related.rds")
SC_related <- readRDS(file = "Relatedness/output_SC_related.rds")

all_related <- rbind(data.frame("ind1" = HG_related$relatedness$ind1.id, "ind2" = HG_related$relatedness$ind2.id, "trioml" = HG_related$relatedness$trioml, "pop" = "HG"),
      data.frame("ind1" = KA_related$relatedness$ind1.id, "ind2" = KA_related$relatedness$ind2.id, "trioml" = KA_related$relatedness$trioml, "pop" = "KA"),
      data.frame("ind1" = nSE_related$relatedness$ind1.id, "ind2" = nSE_related$relatedness$ind2.id, "trioml" = nSE_related$relatedness$trioml, "pop" = "nSE"),
      data.frame("ind1" = SE_related$relatedness$ind1.id, "ind2" = SE_related$relatedness$ind2.id, "trioml" = SE_related$relatedness$trioml, "pop" = "SE"),
      data.frame("ind1" = SC_related$relatedness$ind1.id, "ind2" = SC_related$relatedness$ind2.id, "trioml" = SC_related$relatedness$trioml, "pop" = "SC"))

first_order <- all_related %>% filter(trioml > 0.4)

remove <- unique(first_order$ind1) #remove for admixture analysis
```
# Run admixture population clustering
Admixture script: run_admixture.sh and python wrapper (cite)
```{r}
vcfR_panel_filt_rel <- vcfR_panel_filt[, !colnames(vcfR_panel_filt@gt) %in% remove]

vcfR::write.vcf(vcfR_panel_filt_rel, "data/vcfR_panel_filt_rel.vcf.gz") # Test filt/shared loci
```
## Admixture
#### CV error
```{r figures-side, out.width="30%", echo=FALSE, warning=FALSE}
library(ggplot2)
WR_CV_error <- read.delim("Admixture/panel_loci/related_removed/related_removed.CV_Avg.txt")

png("figures/tissues_manuscript/Supp_CV_error.png", width = 6000, height = 4000, res = 600)
ggplot(WR_CV_error, aes(x=K, y=CV_Avg)) + 
  geom_line() +
  geom_pointrange(aes(ymin=CV_Avg-CV_Stdev, ymax=CV_Avg+CV_Stdev), size = 0.3) +
  labs(title = "") +
  xlab("K-value") + ylab("Cross Validation Error Average") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20))
dev.off()
```
### Admixture barplots
### Read in Admixture results files
```{r}
library(pophelper)

afiles <- list.files("Admixture/panel_loci/related_removed/Outputs-related_removed/", , full.names = T, pattern = "\\.Q$")

alist <- readQ(files=afiles)
```
### Summarize Admixture runs
```{r}
summariseQ(tabulateQ(alist))
```
### Row labels
```{r}
popmap_rel <- popmap_relabeled[ popmap_relabeled$id %in% colnames(vcfR_panel_filt_rel@gt), ]

sites <- data.frame("id" = locations$UI.Barcode, "site" = locations$Region)
sites$site[is.na(sites$site)] <- "Unknown"

popmap_rel <- left_join(popmap_rel, sites)

rownames(alist[[1]]) <- popmap_rel$id

labels <-  data.frame(site = as.character(popmap_rel$site), pop = as.character(popmap_rel$pop))

labels[labels == "British Columbia"] <- "B.C."
```
## View barplots K = 2 through 6
```{r}
# Organize
qlist <- alist[c(11,16,21,26,31)]

rownames(qlist$related_removed.2.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.2.1.Q))
rownames(qlist$related_removed.3.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.3.1.Q))
rownames(qlist$related_removed.4.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.4.1.Q))
rownames(qlist$related_removed.5.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.5.1.Q))
rownames(qlist$related_removed.6.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.6.1.Q))

qlist$related_removed.4.1.Q <- data.frame("Cluster1" = qlist$related_removed.4.1.Q$Cluster3, "Cluster2" = qlist$related_removed.4.1.Q$Cluster2, "Cluster3" = qlist$related_removed.4.1.Q$Cluster4, "Cluster4" = qlist$related_removed.4.1.Q$Cluster1)

qlist$related_removed.5.1.Q <- data.frame("Cluster1" = qlist$related_removed.5.1.Q$Cluster1, "Cluster2" = qlist$related_removed.5.1.Q$Cluster4, "Cluster3" = qlist$related_removed.5.1.Q$Cluster5, "Cluster4" = qlist$related_removed.5.1.Q$Cluster2, "Cluster5" = qlist$related_removed.5.1.Q$Cluster3)

arrange(qlist$related_removed.6.1.Q, desc(Cluster6))

qlist$related_removed.6.1.Q <- data.frame("Cluster1" = qlist$related_removed.6.1.Q$Cluster6, "Cluster2" = qlist$related_removed.6.1.Q$Cluster3, "Cluster3" = qlist$related_removed.6.1.Q$Cluster4, "Cluster4" = qlist$related_removed.6.1.Q$Cluster2, "Cluster5" = qlist$related_removed.6.1.Q$Cluster1, "Cluster6" = qlist$related_removed.6.1.Q$Cluster5)
```
### Plot
```{r}
# Define your custom color palette
#"#fce5c9","#b5dee4","#3a7380","#f6b26b","#b45f06","#682502","#431801"

my_colors <- c("#6aa8b1","#f9d5a9","#431801","#eca255","#b45f06","#783513")  # Replace/add colors as needed for your K

# Plot with custom colors
p <- plotQ(
  qlist = qlist,
  imgoutput = "join",
  returnplot = TRUE,
  exportplot = FALSE,
  basesize = 10,
  grplab = labels,
  ordergrp = TRUE,
  subsetgrp = c("Sitkalidak Island", "Uganik Island", "Kodiak Island", "Whale Island", "Afognak Island", "Hinchinbrook Island", "Hawkins Island", "Shelter Island", "Mainland Juneau", "Douglas Island", "Admiralty Island", "Chichagof Island", "Baranof Island", "Krestof Island", "Partofshikof Island", "Kruzof Island", "Unknown", "Mitkof Island", "Kupreanof Island", "Cleveland Peninsula", "Revillagigedo Island", "Gravina Island", "Prince of Wales Island", "B.C.", "Graham Island"),
  grplabangle = 0,
  grplabsize = 4,
  linesize = 0,
  pointsize = 10,
  grplabpos = 0.5,
  grplabspacer = 0.5,
  grplabjust = 0.5,
  grplabheight = 30,
  splabsize = 28,
  divgrp = "pop",
  divsize = 1,
  divtype = 1,
  clustercol = my_colors   # <--- custom color palette here
)

#c("Kodiak", "Southcentral", "northern Southeast", "Unknown", "middle Southeast", "southern Southeast", "B.C.", "Haida Gwaii")
png("figures/tissues_manuscript/SBTD_admixture.png", width = 6000, height = 6000, res = 600)
grid.arrange(p$plot[[1]])
dev.off()
```
# Locus variability and utility by population
## Format data
```{r}
popmap_var <- data.frame(id = popmap_relabeled$id, pop = as.character(popmap_relabeled$pop))
## Combine BC with SE since low sample size and groups as a population, put NA samples in the  SE/BC group
popmap_var$pop[popmap_var$pop == "British Columbia"] <- "P. of Wales Is. & B.C."
popmap_var$pop[popmap_var$pop == "P. of Wales Is."] <- "P. of Wales Is. & B.C."
popmap_var$pop[popmap_var$pop == "Unknown"] <- "northern Southeast"

genind <- vcfR2genind(vcfR_panel_filt)
pop(genind) <- popmap_var$pop # Pop by Island/site or Region
genind_seppop <- seppop(genind)
```
## Locus stats by region
```{r}
pop_summary <- map_dfr(
  names(genind_seppop),
  function(pop_name) {
    g <- genind_seppop[[pop_name]]
    summ <- summary(g)
    maf <- minorAllele(g)
    
    tibble(
      Broad_region = pop_name,
      Locus = locNames(g),
      n = summ$n,
      Ho = summ$Hobs,
      MAF = as.numeric(maf),
      label = paste(pop_name," n=", summ$n, sep = "")
    )
  }
)

head(pop_summary)
```
### Boxplot of observed het and maf for panel loci by region
```{r}
color_pal <- c("#431801", "#783513","#b45f06","#eca255","#3a7380","#9ac9d0","#f9d5a9")

pop_summary$Broad_region <- factor(pop_summary$Broad_region, levels = c("Haida Gwaii", "P. of Wales Is. & B.C.","southern Southeast", "middle Southeast", "northern Southeast", "Southcentral", "Kodiak"))
pop_summary$label <- factor(pop_summary$label, levels = unique(pop_summary$label[order(pop_summary$Broad_region)]))

library(scales)  # for label_number()

# p1: Observed Heterozygosity
p1 <- ggplot(pop_summary, aes(x = label, y = Ho, fill = Broad_region)) +
  geom_boxplot(alpha = 0.9, color = "black") +  # 90% opacity, black outline
  scale_fill_manual(values = color_pal, guide = "none") +  # apply palette to fill
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = NULL,
    y = "Observed Heterozygosity",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 12)
  )

# p2: Minor Allele Frequency

p2 <- ggplot(pop_summary, aes(x = label, y = MAF, fill = Broad_region)) +
  geom_boxplot(alpha = 0.9, color = "black") +  # 90% opacity, black outline
  scale_fill_manual(values = color_pal, guide = "none") +  # apply palette to fill
  scale_x_discrete(labels = function(x) str_wrap(x, width = 15)) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = "Region",
    y = "Minor Allele Frequency"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(size = 12)
  )
```
# Save boxplot
```{r}
png("figures/tissues_manuscript/SBTD_panel_het_maf_region_boxplots.png", width = 5000, height = 3000, res = 600)

grid.arrange(p1, p2, ncol = 1, heights = c(1,1.6))

dev.off()
```
### Number variable by region
```{r}
invar_summary_df <- pop_summary %>%
  filter(MAF == 0) %>% # Invariable loci
  group_by(Broad_region) %>%       # group by region
  summarise(n_invar_loci = n()) %>%      # count number of loci
  arrange(desc(n_invar_loci))           # sort by count

invar_summary_df

var_summary_df <- pop_summary %>%
  filter(Ho > 0.2) %>%
  filter(MAF > 0.2) %>% # Invariable loci
  group_by(Broad_region) %>%       # group by region
  summarise(n_loci = n()) %>%      # count number of loci
  arrange(n_loci)           # sort by count

var_summary_df

left_join(var_summary_df, invar_summary_df)
```
# PID/PIDSib
## Calculate PID and PIDsib by population/region
Original simulation script from Dana Morin, modified for multiple regions/populations
```{r}
library(dplyr)
library(tidyr)
library(purrr)

# Parameters
N <- 1000 # number of simulations
x <- 40   # max number of loci to simulate through

# Function to calculate PID and PIDsib for one population
calc_pid <- function(genind_obj, popname, N, x) {
  
  # allele frequencies
  freq <- makefreq(genind_obj) %>% as.data.frame()
  df <- colMeans(freq, na.rm = TRUE) %>% as.data.frame()
  row_names <- rownames(df)
  
  locus_names <- sub("^(([^:]+):[^:]+):.*$", "\\1", row_names)
  
  df$Marker <- locus_names
  df$allele <- ifelse(grepl("\\.0$", row_names), "X1", "X2")
  
  allele_freqs <- pivot_wider(df, 
                              id_cols = Marker,
                              names_from = allele, 
                              values_from = .)
  
  # format allele frequencies into list
  freqs <- allele_freqs
  al.freq.ls <- NULL
  for (i in 1:length(freqs$Marker)) {
    addon <- unname(unlist(c(freqs[i, -1])))
    addon <- addon[!is.na(addon)]
    addon <- list(addon)
    al.freq.ls <- c(al.freq.ls, addon)
  }
  
  # simulation
  PID.pop.tbl <- matrix(NA, nrow = x, ncol = 2)
  PID.pop.sib.tbl <- matrix(NA, nrow = x, ncol = 2)
  
  for (j in 1:x) { 
    PIpop.loci <- NULL
    PIpop.sib.loci <- NULL
    
    for (h in 1:N) {
      samp <- sample(al.freq.ls, j, replace = FALSE)
      PIpop <- NULL
      PIpop.sib <- NULL
      
      for (i in 1:length(samp)) {
        r1 <- sum(samp[[i]]^4)
        r2 <- outer(X = samp[[i]], Y = samp[[i]], FUN = function(X,Y)(2*X*Y)^2)
        r <- sum(r2[lower.tri(r2)]) + r1
        s <- 0.25 + (0.5 * sum(samp[[i]]^2)) + (0.5 * sum(samp[[i]]^2)^2) -
          (0.25 * sum(samp[[i]]^4))
        
        PIpop <- c(PIpop, r)
        PIpop.sib <- c(PIpop.sib, s)
      }
      PIpop.loci <- c(PIpop.loci, prod(PIpop))
      PIpop.sib.loci <- c(PIpop.sib.loci, prod(PIpop.sib))
    }
    
    PID.pop.tbl[j, 1] <- j
    PID.pop.tbl[j, 2] <- mean(PIpop.loci)
    PID.pop.sib.tbl[j, 1] <- j
    PID.pop.sib.tbl[j, 2] <- mean(PIpop.sib.loci)
  }
  
  PID.pop.tbl <- as.data.frame(PID.pop.tbl)
  colnames(PID.pop.tbl) <- c("Loci", "Avg.PID")
  
  PID.pop.sib.tbl <- as.data.frame(PID.pop.sib.tbl)
  colnames(PID.pop.sib.tbl) <- c("Loci", "Avg.PIDsib")
  
  # merge PID and PIDsib, add population column
  all_PID <- left_join(PID.pop.sib.tbl, PID.pop.tbl, by = "Loci") %>%
    mutate(Population = popname)
  
  return(all_PID)
}

# ---- Run across all populations ----
all_PID <- map2_dfr(genind_seppop, names(genind_seppop),
                    ~calc_pid(.x, .y, N, x))

# Final output
head(all_PID)
```
# Plot PID and PIDsib
```{r}
pid_color_pal <- c("#431801", "#783513","#b45f06","#eca255","#3a7380","#9ac9d0","#f9d5a9")

all_PID$Population <- factor(all_PID$Population, levels = c("Haida Gwaii", "P. of Wales Is. & B.C.","southern Southeast", "middle Southeast", "northern Southeast", "Southcentral", "Kodiak")) 


p_PID <- ggplot(all_PID, aes(x = Loci, y = Avg.PID, color = Population)) +
  geom_line(size = 1) +
  geom_hline(aes(yintercept = 0.001, linetype = "PID = 0.001"), color = "black") +
  scale_linetype_manual(
    values = c("PID = 0.001" = "dashed")
  ) +
  scale_x_continuous(limits = c(0, 30)
  ) +
  coord_cartesian(ylim = c(0, 0.1)
  ) +
  scale_color_manual(values = pid_color_pal) +
  labs(
  #  title = "PID with Panel, Nsim = 1000",
    x = "Number of Loci",
    y = "Probability of Identity",
    color = NULL,
    linetype = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "none")

p_PIDsib <- ggplot(all_PID, aes(x = Loci, y = Avg.PIDsib, color = Population)) +
  geom_line(size = 1.5) +
  geom_hline(aes(yintercept = 0.001, linetype = "PIDsib = 0.001"), color = "black") +
  scale_linetype_manual(
    values = c("PIDsib = 0.001" = "dashed")
  ) +
  scale_x_continuous(limits = c(0, 40)
  ) +
  coord_cartesian(ylim = c(0, 0.1)
  ) +
  scale_color_manual(values = pid_color_pal) +
  labs(
    #title = "PIDsib with Panel, Nsim = 1000",
    x = "Number of Loci",
    y = "Probability of Identity",
    color = NULL,
    linetype = NULL
  ) +
  theme_minimal(base_size = 14)
```
### Save plots
```{r}
png("figures/tissues_manuscript/SBTD_PIDsib.png", width = 5000, height = 3000, res = 600)

p_PIDsib

dev.off()

png("figures/tissues_manuscript/SBTD_PID.png", width = 5000, height = 3000, res = 600)

p_PID

dev.off()
```
# List of variable loci by region
```{r}
# Assess if any loci are out of HWE
HWE_SC <- data.frame(hw.test(genind_seppop$Southcentral)) 
HWE_nSE <- data.frame(hw.test(genind_seppop$`northern Southeast`))
HWE_HG <- data.frame(hw.test(genind_seppop$`Haida Gwaii`))
HWE_KA <- data.frame(hw.test(genind_seppop$Kodiak))
HWE_POW_BC <- data.frame(hw.test(genind_seppop$`P. of Wales Is. & B.C.`))
HWE_mSE <- data.frame(hw.test(genind_seppop$`middle Southeast`))
HWE_sSE <- data.frame(hw.test(genind_seppop$`middle Southeast`))

# P adjust with Benjamini Hochberg FDR method
HWE_SC_subset <- HWE_SC[p.adjust(HWE_SC$Pr.exact, method = "BH") < 0.05, ] # none out of HWE
HWE_nSE_subset <- HWE_nSE[p.adjust(HWE_nSE$Pr.exact, method = "BH") < 0.05, ]
HWE_HG_subset <- HWE_HG[p.adjust(HWE_HG$Pr.exact, method = "BH") < 0.05, ] # none out of HWE
HWE_KA_subset <- HWE_KA[p.adjust(HWE_KA$Pr.exact, method = "BH") < 0.05, ]
HWE_POW_BC_subset <- HWE_POW_BC[p.adjust(HWE_POW_BC$Pr.exact, method = "BH") < 0.05, ]  # none out of HWE
HWE_mSE_subset <- HWE_mSE[p.adjust(HWE_mSE$Pr.exact, method = "BH") < 0.05, ]
HWE_sSE_subset <- HWE_sSE[p.adjust(HWE_sSE$Pr.exact, method = "BH") < 0.05, ]

HWE_nSE_subset$Region <- "nSE"
HWE_KA_subset$Region <- "KA"
HWE_mSE_subset$Region <- "mSE"
HWE_sSE_subset$Region <- "sSE"

HWE <- data.frame(Locus = c(rownames(HWE_nSE_subset), rownames(HWE_KA_subset), rownames(HWE_mSE_subset), rownames(HWE_sSE_subset)), info = c(HWE_nSE_subset$Region, HWE_KA_subset$Region, HWE_mSE_subset$Region, HWE_sSE_subset$Region))

High_het <- pop_summary %>% filter(Ho > 0.7) %>% select(Locus) %>% unique()
High_het$info <- "High_het"

HWE_high_het <- rbind(HWE, High_het) %>% arrange(Locus)

unique(HWE_high_het$Locus) # remove these from recommended loci

var_loci_df <- pop_summary %>%
  filter(Ho > 0.2) %>%
  filter(MAF > 0.2) %>% # Invariable loci
  group_by(Broad_region)      # group by region

recommended <- var_loci_df %>% filter(!Locus %in% unique(HWE_high_het$Locus))
```
# Create recommended loci table
```{r}
F_R_seq <- read.csv("data/F_R_sequences.csv")
F_R_seq$CHROM_POS <- paste(F_R_seq$CHROM,F_R_seq$POS, sep = "_")

locus_info <- data.frame(vcfR_panel_filt@fix) %>% select(CHROM:ALT)

rec_locus_info <- locus_info[locus_info$ID %in% unique(recommended$Locus), ]
rec_locus_info$CHROM_POS <- paste(rec_locus_info$CHROM, rec_locus_info$POS, sep = "_")

rec_F_R <- left_join(rec_locus_info, F_R_seq, by = "CHROM_POS")
recommended_table <- left_join(recommended, rec_F_R, by = c("Locus"="ID")) %>% select(Broad_region:MAF,CHROM.x:ALT, Forward.sequence, Reverse.sequence)

colnames(recommended_table) <- c("Broad Region", "Locus ID", "N Samples", "Observed Heterozygosity", "Minor Allele Frequency", "Chromosome", "Position", "Reference Allele", "Alternate Allele", "Forward Sequence", "Reverse Sequence")

write.csv(recommended_table, file = "output/recommended_locus_set_by_region.csv")
```
