---
title: "SBTD_Tissue"
author: "Elise Stacy"
date: "2025-08-26"
output: html_document
---
# Load packages
```{r}
# Additional filtering
library(vcfR)
library(SNPfiltR) 

library(adegenet) # Locus div calcs
library(pegas) # For HWE calcs

# Data structure manipulation
library(stringr)
library(purrr)
library(dplyr)
library(tidyr)
library(tibble)

# Plotting
library(ggplot2)
library(gridExtra)

# Maps
library(ggspatial)
library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
```
## Load error calculation functions (code slightly modified from https://github.com/arianacerreta)
```{r}
## Scoring matching genotypes
#make a function
score_gt_match<-function(gt1,gt2){
  #Handle missing genotypes
  if (is.na(gt1) & is.na(gt2)) return("BothNA")
  else if (is.na(gt1) | is.na(gt2)) return("OneNA")
  
  #split alleles
  a1 <- unlist(strsplit(gt1, "[/|]"))
  a2 <- unlist(strsplit(gt2, "[/|]"))
  
  #sort alleles for unphased comparison
  a1<- sort(a1)
  a2<- sort(a2)
  
  #score
  if (identical(a1,a2)) return("Full Match")
  else if (any(a1 %in% a2)) return("Partial Match")
  else return("Mismatch")
}

format_paired_gt_df <- function(vcfR_samples, popmap_samples) {

  # Extract base replicate IDs
  popmap_samples <- popmap_samples %>%
    mutate(ReplicateID = sub(".*_", "", id))

  # Keep only replicate groups with >=2 samples
  replicate_counts <- popmap_samples %>%
    count(ReplicateID) %>%
    filter(n >= 2)

  # Filter popmap_samples to those with replicates
  popmap_samples <- popmap_samples %>%
    filter(ReplicateID %in% replicate_counts$ReplicateID)

  # Subset VCF to matching sample names
  # Ensure only selected samples are kept
  sample_ids <- popmap_samples$id
  vcfR_samples@gt <- vcfR_samples@gt[, c("FORMAT", sample_ids)]


  # Extract GT matrix
  gt_df <- extract.gt(vcfR_samples, element = "GT", as.numeric = FALSE) %>% as.data.frame()
  gt_df$CHROM <- getCHROM(vcfR_samples)
  gt_df$POS <- as.numeric(getPOS(vcfR_samples))
  gt_df <- gt_df %>% relocate(CHROM, POS)

  # Pivot to long format
  gt_df <- pivot_longer(gt_df, cols = -c(CHROM, POS), names_to = "SampleID", values_to = "GT")
  gt_df <- gt_df %>%
    mutate(ReplicateID = sub(".*_", "", SampleID))

  # Generate all pairwise combinations of replicates
  replicate_map <- popmap_samples %>%
    select(ReplicateID, id) %>%
    group_by(ReplicateID) %>%
    filter(n() >= 2) %>%
    summarise(pair_df = list(as.data.frame(t(combn(id, 2)))), .groups = "drop") %>%
    unnest(pair_df) %>%
    rename(Sample1 = V1, Sample2 = V2)


  # Join with genotypes
  paired_gt_df <- replicate_map %>%
    left_join(gt_df, by = c("Sample1" = "SampleID")) %>%
    rename(GT1 = GT, CHROM = CHROM, POS = POS) %>%
    left_join(gt_df, by = c("Sample2" = "SampleID", "CHROM", "POS")) %>%
    rename(GT2 = GT)

  return(paired_gt_df)
}

# Process data and create summaries in one function
calculate_gt_pairs_errors <- function(paired_gt_df) {
  # Create gt_pairs_mismatch dataframe
  gt_pairs_mismatch <- paired_gt_df %>%
    rowwise() %>%
    mutate(MatchType = score_gt_match(GT1, GT2)) %>%
    ungroup()
  
  # Calculate summary_by_pair
  summary_by_pair <- gt_pairs_mismatch %>%
    group_by(Sample1, Sample2) %>%
    summarize(
      N = n(),
      Full_Match = sum(MatchType == "Full Match", na.rm = TRUE),
      Partial_Match = sum(MatchType == "Partial Match", na.rm = TRUE),
      Mismatch = sum(MatchType == "Mismatch", na.rm = TRUE),
      Missing_Full = sum(MatchType == "BothNA", na.rm = TRUE),
      Missing_Partial = sum(MatchType == "OneNA", na.rm = TRUE),
      Full_Match_Rate = Full_Match / (Full_Match + Partial_Match + Mismatch),
      Mismatch_Rate = (Mismatch + Partial_Match) / (Full_Match + Partial_Match + Mismatch),
      Full_Missingness = Missing_Full / N,
      Partial_Missingness = Missing_Partial / N,
      Inclusive_Missingness = (Missing_Full + Missing_Partial) / N
    )
  
  # Calculate summary_by_locus
  summary_by_locus <- gt_pairs_mismatch %>%
    group_by(CHROM, POS) %>%
    summarize(
      N=n(),
      Full_Match = sum(MatchType == "Full Match", na.rm = TRUE),
      Mismatch = sum(MatchType == "Mismatch", na.rm = TRUE),
      Partial_Match = sum(MatchType == "Partial Match", na.rm = TRUE),
      Missing_Full = sum(MatchType == "BothNA", na.rm = TRUE),
      Missing_Partial = sum(MatchType == "OneNA", na.rm = TRUE),
      Full_Match_Rate = Full_Match / (Full_Match + Partial_Match + Mismatch),
      Both_Mismatch_Rate = (Partial_Match + Mismatch) / (Full_Match + Partial_Match + Mismatch),
      Full_Missingness = Missing_Full / N,
      Partial_Missingness = Missing_Partial / N,
      Inclusive_Missingness = (Missing_Full + Missing_Partial) / N
    )
  
  # Return both summaries
  list(summary_by_pair = summary_by_pair,
       summary_by_locus = summary_by_locus)
}
```
# Read in sample data
```{r}
snps_vcfR <- read.vcfR("/mnt/ceph/estacy/SBTDeer/Tissue_samples/05_filter/populations_R0.5_mac2/populations.snps.vcf")

locations <- read.csv("/mnt/ceph/estacy/SBTDeer/R_analysis/data/Tissue_locations.csv", na.strings = c("","NA"))

locations <- locations %>%
  mutate(
    Populations = case_when(
      Region %in% c("Gravina Island",
                    "Revillagigedo Island",
                    "Cleveland Peninsula") ~ "SSEAK",
      Region %in% c("Kupreanof Island", "Mitkof Island") ~ "CSEAK",
      Region == "Prince of Wales Island" ~ "P. of Wales Is.",
      Broad_region == "northern Southeast" ~ "NSEAK",
      Broad_region == "Kodiak Archipelago" ~ "Kodiak",
      Broad_region == "Southcentral" ~ "SCAK",
      Broad_region == "Haida Gwaii" ~ "Haida Gwaii",
      TRUE ~ "Unknown"
    )
  )
```

# Subset (whitelist) to designed panel loci
```{r}
panel_snps <- read.csv("/mnt/ceph/estacy/SBTDeer/R_analysis/data/Eacker_Deer_248_Loci_Names.csv", header = F)

# Whitelist snps panel was designed for only
panel_snps <- panel_snps %>%
  separate(V1, into = c("CHROM", "POS"), sep = "_")

# Create combined keys
vcf_keys <- paste(snps_vcfR@fix[, "CHROM"], snps_vcfR@fix[, "POS"], sep = "_")
panel_keys <- paste(panel_snps$CHROM, panel_snps$POS, sep = "_")

# Filter based on matching keys
vcfR_panel <- snps_vcfR[vcf_keys %in% panel_keys, ]
```
# Filtering to compare schemes
## Error rates in replicates
```{r}
vcfR <- vcfR_panel

# Initialize storage data frames
pair_errors <- data.frame()
locus_errors <- data.frame()

# Loop through depth values
for (depth in seq(5,30, by = 5)) {
  # Loop through missingness from 80% through fully missing
    for (missing in seq(0.1,0.4, by = 0.05)) {
    
    # Apply hard filter with current depth and GQ
    vcfR_filtered <- hard_filter(vcfR = vcfR, depth = depth, gq = 30)
    vcfR_missing <- missing_by_sample(vcfR = vcfR_filtered, cutoff = missing)
    
    popmap_samples <- data.frame(id=colnames(vcfR_missing@gt)[2:length(colnames(vcfR_missing@gt))])

    # Format paired genotype data frame
    paired_gt_df <- format_paired_gt_df(vcfR_missing, popmap_samples)
    
    # Calculate error rate
    error_rate <- calculate_gt_pairs_errors(paired_gt_df)
    
    # Add current Depth and GQ values to error summaries
    pair_error <- cbind(error_rate$summary_by_pair, Depth = depth, Missing = missing)
    locus_error <- cbind(error_rate$summary_by_locus, Depth = depth, Missing = missing)
    
    # Accumulate error data
    pair_errors <- rbind(pair_errors, pair_error)
    locus_errors <- rbind(locus_errors, locus_error)
    }
}
```
# Visualize error rate by filtering scheme
```{r}
pair_errors$ID_1 <- sub(".*_", "", pair_errors$Sample1)
pair_errors$ID_2 <- sub(".*_", "", pair_errors$Sample2) 

pair_err_summ <- pair_errors %>% # Summarize mismatch data
  group_by(Depth, Missing) %>%
  summarise(
    N_Samples = n(),
    Avg_Mismatch = round(mean(Mismatch_Rate, na.rm = TRUE), 4),
    Max_Mismatch = round(max(Mismatch_Rate, na.rm = TRUE), 4),
    Min_Locus_Overlap = min(((1 - Inclusive_Missingness) * N), na.rm = TRUE),
    Avg_Missing = round(mean(Missing_Full, na.rm = TRUE), 1)
  )

png("figures/tissues_manuscript/Supp_error_rates.png", width = 3000, height = 2000, res = 600)
# Plot error rates
ggplot(pair_err_summ, aes(x=Missing, y = Depth, fill = Avg_Mismatch)) +
  geom_tile() +
  scale_fill_gradient2(low = "lightblue", mid ="purple", high = "red", midpoint = 0.0025) +
  geom_text(aes(label = N_Samples), color = "black", size = 3) +
  labs(title = "Replicate Tissue Retained and Average mismatch", subtitle = "Across missingness and depth filtering") +
  ylab("Minimum Genotype Depth") +
  xlab("Sample Missingness")
dev.off()

pair_errors %>% filter(Depth == 10 & Missing == 0.4) %>% arrange(desc(Mismatch_Rate))
```
# Select one replicate per lab_id with the lowest missing_data and highest depth
```{r}
# Data frame of full unique IDs and lab IDs
sample_ids <- data.frame(id = sub(".*_", "", colnames(vcfR_panel@gt)[-1]), full_id = colnames(vcfR_panel@gt)[-1])
# Create set of replicate only ids
replicates <- sample_ids %>%
  group_by(id) %>%
  filter(n() > 1)
# Create vectors of replicate vs not rep samples
replicate_samples <- intersect(colnames(vcfR_panel@gt)[-1], replicates$full_id)
non_rep_samples <- setdiff(colnames(vcfR_panel@gt)[-1], replicates$full_id)

# Seperate vcfR objects in replicated samples and non rep objects
vcfR_reps <- vcfR_panel[samples = replicate_samples]
vcfR_non_reps <- vcfR_panel[samples = non_rep_samples]

# Apply locus filters
vcfR_reps_filt <- hard_filter(vcfR_reps, gq = 30, depth = 10)
vcfR_reps_filt <- min_mac(vcfR_reps_filt, min.mac = 2)

# Calculate missingness and depth by sample
missing_df_reps <- missing_by_sample(vcfR_reps_filt)
rep_stats <- missing_df_reps$unfiltered.stats
rep_stats$id <- sub(".*_", "", rep_stats$samples) # Format base sample ID only column

best_replicate <- rep_stats %>%
  group_by(id) %>%
  slice_min(prop.missing, with_ties = TRUE) %>% # Choose replicate with least missing data, allowing ties
  slice_max(mean.depth, with_ties = FALSE) %>% # Only paired should be ties now, choose replicate with greatest depth
  ungroup()

popmap_unique <- data.frame(id = c(best_replicate$samples, non_rep_samples))

vcfR_unique <- vcfR_panel[samples = popmap_unique$id]
```
# Panel locus filtering
```{r}
vcfR_dp <- hard_filter(vcfR_unique, gq = 30, depth = 10)
vcfR_missing <- missing_by_sample(vcfR_dp, cutoff = 0.2)
vcfR_mac <- min_mac(vcfR_missing, min.mac = 2)
vcfR_snp_miss <- missing_by_snp(vcfR_mac, cutoff = 0.5)

vcfR_panel_filt <- vcfR_snp_miss
```
### Mean depth
```{r}
dp <- extract.gt(vcfR_panel_filt, element = "DP", as.numeric=TRUE)

colMeans(dp, na.rm = T) %>% mean(na.rm = T)
colMeans(dp, na.rm = T) %>% min()
colMeans(dp, na.rm = T) %>% max()
```
# Format popmap
```{r}
popmap <- data.frame(id=colnames(vcfR_panel_filt@gt)[-1])

popmap_relabeled <- data.frame(id = sub(".*_", "", popmap$id))

sample_locations <- left_join(popmap_relabeled, locations, by = c("id"="UI.Barcode"))

popmap_relabeled$pop <- sample_locations$Populations

popmap_relabeled$pop <- factor(popmap_relabeled$pop, levels = c("Kodiak", "SCAK", "NSEAK", "CSEAK", "SSEAK", "P. of Wales Is.", "Haida Gwaii", "Unknown"))

colnames(vcfR_panel_filt@gt) <- c("FORMAT", popmap_relabeled$id)
```
## Map
### Format data to map
```{r}
locations$Populations <- factor(locations$Populations, levels = c("Kodiak", "SCAK", "NSEAK", "CSEAK", "SSEAK", "P. of Wales Is.", "Haida Gwaii", "Unknown"))

# Sample size by location

locations_clean <- locations %>% filter(!is.na(Lon)) %>% filter(!is.na(UI.Barcode))

locations_clean <- locations_clean %>%
  group_by(Populations) %>%
  mutate(Population_n = paste(Populations," (n=", n(), ")", sep = "")) %>%
  ungroup()

SBTD_sites <- locations_clean[locations_clean$UI.Barcode %in% popmap_relabeled$id, ] # Plot samples with locations that passed filtering

labels_vector <- setNames(
  unique(SBTD_sites$Population_n),
  unique(SBTD_sites$Populations))

populations_color_pal <- c("#f9d5a9","#b5dee4","#3a7380","#eca255","#b45f06","#863e19","#682502","#431801","grey")
```
# Plot and save map of samples
```{r}
world_sf <- ne_countries(
  scale = "medium",
  returnclass = "sf"
)

locations_sf <- st_as_sf(
  locations_clean,
  coords = c("Lon", "Lat"),
  crs = 4326
)

world_sf <- st_transform(world_sf, 3338)      # Alaska Albers
locations_sf <- st_transform(locations_sf, 3338)

bbox_ll <- st_bbox(
  c(xmin = -156, ymin = 54.5, xmax = -125, ymax = 59.5),
  crs = st_crs(4326)
)

bbox_proj <- st_transform(st_as_sfc(bbox_ll), 3338)

png("figures/tissues_manuscript/SBTD_sites_map.png",
    width = 6000, height = 6000, res = 600)

ggplot() +
  geom_sf(data = world_sf,
          fill = "grey98",
          color = "grey50") +
  geom_sf(data = locations_sf,
          aes(fill = Populations),
          size = 4,
          shape = 21,
          stroke = 0.2) +
  scale_fill_manual(
    values = populations_color_pal,
    labels = labels_vector,
    guide = guide_legend(ncol = 1)
  ) +
  coord_sf(
    crs = st_crs(3338),
    xlim = c(st_bbox(bbox_proj)["xmin"], st_bbox(bbox_proj)["xmax"]),
    ylim = c(st_bbox(bbox_proj)["ymin"], st_bbox(bbox_proj)["ymax"]),
    expand = FALSE
  ) +
  annotation_scale(
    location = "bl",
    unit_category = "metric",
    width_hint = 0.25,
    text_cex = 0.9
  ) +
  annotation_north_arrow(
    location = "tl",
    which_north = "true",
    style = north_arrow_orienteering
  ) +
  theme_minimal() +
  labs(x = NULL, y = NULL) +
  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.text = element_text(size = 16),
    axis.text = element_text(size = 16)
  )

dev.off()
```



# Plot region popout map
```{r}
# Get USA states
states <- ne_states(country = "United States of America", returnclass = "sf")

# Get Canada provinces/territories (Admin-1)
canada <- ne_states(country = "Canada", returnclass = "sf")

# Combine for labeling
regions <- rbind(
  states %>% select(name, geometry),
  canada %>% select(name, geometry)
)

# Compute label positions
regions_centroids <- st_centroid(regions)

regions_centroids <- regions_centroids %>%
  mutate(
    label_x = st_coordinates(geometry)[,1],
    label_y = st_coordinates(geometry)[,2]
  )

# Update label format and locations
regions_centroids <- regions_centroids %>%
  mutate(
    label_x = ifelse(name == "British Columbia", label_x - 7, label_x),
    label_y = ifelse(name == "British Columbia", label_y - 4.5, label_y),    
    label_x = ifelse(name == "Yukon", label_x - 1.5, label_x),
    label_y = ifelse(name == "Yukon", label_y + .5, label_y),  
    label_name = case_when(
      name == "British Columbia" ~ "B.C.\nCAN",
      name == "Yukon" ~ "Yukon\nCAN",
      name == "Alaska" ~ "Alaska\nUSA",
      TRUE ~ name
    )
  ) %>%
  filter(name != "Northwest Territories")

png("figures/tissues_manuscript/region_map.png", width = 6000, height = 6000, res = 600)

ggplot() +
  geom_sf(data = ne_countries(scale = 50, returnclass = "sf"),
          fill = "grey", color = "grey70") +
  geom_sf(data = states, fill = NA, color = "#444444", size = 1.5) +
  geom_sf(data = canada, fill = NA, color = "#444444", size = 1.5) +
  geom_text(
    data = regions_centroids,
    aes(x = label_x, y = label_y, label = label_name),
    size = 14,
    fontface = "bold",
    color = "black"
  ) +
  coord_sf(xlim = c(-168.5, -123), ylim = c(48.25, 71), expand = FALSE) +
  theme_void() +  # <<< removes all background lines
  theme(
    text = element_text(size = 22)
  )

dev.off()
```
# PCA
```{r}
pca <- assess_missing_data_pca(vcfR=vcfR_panel_filt, popmap = popmap_relabeled, clustering = FALSE)
```
## Format for region labeled PCA figure
### Add sample size labels
```{r}
# Population order (controls legend order)
pop_order <- c(
  "Kodiak",
  "SCAK",
  "NSEAK",
  "CSEAK",
  "SSEAK",
  "P. of Wales Is.",
  "Haida Gwaii",
  "Unknown"
)

pca$pop <- factor(pca$pop, levels = pop_order)

# Create population sample-size labels
pop_label_df <- pca %>%
  count(pop, name = "n") %>%
  mutate(label = paste0(pop, " (n=", n, ")"))

# Named vector of labels in pop_order
pop_labels_pca <- setNames(
  pop_label_df$label[match(pop_order, pop_label_df$pop)],
  pop_order
)

# Shape palette (named)
populations_shape_pal <- c(
  "Kodiak" = 21,
  "SCAK" = 22,
  "NSEAK" = 21,
  "CSEAK" = 24,
  "SSEAK" = 22,
  "P. of Wales Is." = 23,
  "Haida Gwaii" = 21,
  "Unknown" = 23
)

# Color palette (named)
populations_color_pal_pca <- c(
  "Kodiak" = "#f9d5a9",
  "SCAK" = "#b5dee4",
  "NSEAK" = "#3a7380",
  "CSEAK" = "#eca255",
  "SSEAK" = "#b45f06",
  "P. of Wales Is." = "#863e19",
  "Haida Gwaii" = "#431801",
  "Unknown" = "grey"
)
```
# Plot and save PCAs
```{r}
png("figures/tissues_manuscript/SBTD_pca.png", width = 4500, height = 6000, res = 600)

ggplot() +
  # All populations except BC & Unknown
  geom_point(
    data = subset(pca, !pop %in% c("Unknown")),
    aes(x = PC1, y = PC2, fill = pop, shape = pop),
    size = 10, stroke = 0.5, alpha = 0.9
  ) +

  # BC & Unknown plotted on top
  geom_point(
    data = subset(pca, pop %in% c("Unknown")),
    aes(x = PC1, y = PC2, fill = pop, shape = pop),
    size = 10, stroke = 0.5, alpha = 0.9
  ) +

  scale_fill_manual(
    values = populations_color_pal_pca,
    labels = pop_labels_pca,
    breaks = pop_order
  ) +

  scale_shape_manual(
    values = populations_shape_pal,
    labels = pop_labels_pca,
    breaks = pop_order
  ) +

  theme_light() +
  guides(
    fill  = guide_legend(order = 1, ncol = 2),
    shape = guide_legend(order = 1, ncol = 2)
  ) +

  theme(
    legend.position = "bottom",
    legend.box = "vertical",
    legend.title = element_blank(),
    legend.key.size = unit(0.1, "cm"),
    legend.text = element_text(size = 24),
    axis.text = element_text(size = 24),
    axis.title = element_text(size = 24)
  ) +

  labs(
    x = "PC1 12.91% Variance Explained",
    y = "PC2 5.29% Variance Explained"
  )

dev.off()
```
### Plot PCA by missingness
```{r}
png("figures/tissues_manuscript/SBTD_pca_missing_region.png", width = 6000, height = 3000, res = 600)

ggplot() +
  # All populations except Unknown
  geom_point(
    data = subset(pca, !pop %in% c("Unknown")),
    aes(x = PC1, y = PC2,
        fill  = missing,
        color = pop),
    size = 5,
    shape = 21,     # <-- critical
    stroke = 1.5,
    alpha = 0.9
  ) +
  # Unknown plotted on top
  geom_point(
    data = subset(pca, pop %in% c("Unknown")),
    aes(x = PC1, y = PC2,
        fill  = missing,
        color = pop),
    size = 5,
    shape = 21,     # <-- critical
    stroke = 1.5,
    alpha = 0.9
  ) +
  scale_color_manual(
    values = populations_color_pal_pca,
    labels = pop_labels_pca,
    name   = "Population"
  ) +
  scale_fill_gradient(
    low  = "lightgrey",
    high = "black",
    name = "Missing data"
  ) +
  theme_light()

dev.off()
```
# Relatedness 
Remove related individuals before Admixture
Running trioml on server, submission script: submit_related.sh and related.R
```{r}
library(related)

data(GenotypeData)
input <- readgenotypedata(GenotypeData)

ext.gt <- extract.gt(vcfR_panel_filt) %>% 
                                 t() %>%
                     as.data.frame() %>%
                     rownames_to_column(var = "ID")

gt_split <- ext.gt %>%
  mutate(across(-ID, ~strsplit(., "/"))) %>%
  unnest_wider(where(is.list), names_sep = "_") %>%
  mutate(across(-ID, as.numeric)) %>% 
  mutate(across(
    -ID,
    ~ case_when(
      is.na(.x) ~ 0,
      .x == 0   ~ 1,
      .x == 1   ~ 2,
      TRUE ~ .x    # keeps any unexpected values unchanged
    )
  ))

# Format one input file per population - running relatedness by population/region
gt_pop <- left_join(popmap_relabeled, gt_split, by = c("id" = "ID"))

gt_nSE <- gt_pop %>% filter(pop == "NSEAK" | pop == "Unknown") %>% select(!pop)
gt_SC <- gt_pop %>% filter(pop == "SCAK") %>% select(!pop)
gt_SE <- gt_pop %>% filter(pop == "CSEAK" | pop == "SSEAK") %>% select(!pop)
gt_KA <- gt_pop %>% filter(pop == "Kodiak") %>% select(!pop)
gt_HG <- gt_pop %>% filter(pop == "Haida Gwaii") %>% select(!pop)

input_nSE <- readgenotypedata(gt_nSE)
input_SC <- readgenotypedata(gt_SC)
input_SE <- readgenotypedata(gt_SE)
input_KA <- readgenotypedata(gt_KA)
input_HG <- readgenotypedata(gt_HG)

saveRDS(input_nSE, "Relatedness/input_nSE.rds")
saveRDS(input_SC, "Relatedness/input_SC.rds")
saveRDS(input_SE, "Relatedness/input_SE.rds")
saveRDS(input_KA, "Relatedness/input_KA.rds")
saveRDS(input_HG, "Relatedness/input_HG.rds")
```
## Relatedness results and filter out first order relatedness
```{r}
HG_related <- readRDS(file = "Relatedness/output_HG_related.rds")
KA_related <- readRDS(file = "Relatedness/output_KA_related.rds")
nSE_related <- readRDS(file = "Relatedness/output_nSE_related.rds")
SE_related <- readRDS(file = "Relatedness/output_SE_related.rds")
SC_related <- readRDS(file = "Relatedness/output_SC_related.rds")

all_related <- rbind(data.frame("ind1" = HG_related$relatedness$ind1.id, "ind2" = HG_related$relatedness$ind2.id, "trioml" = HG_related$relatedness$trioml, "pop" = "HG"),
      data.frame("ind1" = KA_related$relatedness$ind1.id, "ind2" = KA_related$relatedness$ind2.id, "trioml" = KA_related$relatedness$trioml, "pop" = "KA"),
      data.frame("ind1" = nSE_related$relatedness$ind1.id, "ind2" = nSE_related$relatedness$ind2.id, "trioml" = nSE_related$relatedness$trioml, "pop" = "nSE"),
      data.frame("ind1" = SE_related$relatedness$ind1.id, "ind2" = SE_related$relatedness$ind2.id, "trioml" = SE_related$relatedness$trioml, "pop" = "SE"),
      data.frame("ind1" = SC_related$relatedness$ind1.id, "ind2" = SC_related$relatedness$ind2.id, "trioml" = SC_related$relatedness$trioml, "pop" = "SC"))

first_order <- all_related %>% filter(trioml > 0.4)

remove <- unique(first_order$ind1) #remove for admixture analysis
```
# Run admixture population clustering
Admixture script: run_admixture.sh and python wrapper (cite)
```{r}
vcfR_panel_filt_rel <- vcfR_panel_filt[, !colnames(vcfR_panel_filt@gt) %in% remove]

vcfR::write.vcf(vcfR_panel_filt_rel, "data/vcfR_panel_filt_rel.vcf.gz") # Use to run Admixture
```
## Admixture
#### CV error
```{r figures-side, out.width="30%", echo=FALSE, warning=FALSE}
library(ggplot2)
WR_CV_error <- read.delim("Admixture/panel_loci/related_removed/related_removed.CV_Avg.txt")

png("figures/tissues_manuscript/Supp_CV_error.png", width = 6000, height = 4000, res = 600)
ggplot(WR_CV_error, aes(x=K, y=CV_Avg)) + 
  geom_line() +
  geom_pointrange(aes(ymin=CV_Avg-CV_Stdev, ymax=CV_Avg+CV_Stdev), size = 0.3) +
  labs(title = "") +
  xlab("K-value") + ylab("Cross Validation Error Average") +
  scale_x_continuous(breaks = scales::pretty_breaks(n = 10)) +
  theme(axis.text = element_text(size = 20), axis.title = element_text(size = 20))
dev.off()
```
### Admixture barplots
### Read in Admixture results files
```{r}
library(pophelper)

afiles <- list.files("Admixture/panel_loci/related_removed/Outputs-related_removed/", , full.names = T, pattern = "\\.Q$")

alist <- readQ(files=afiles)
```
### Summarize Admixture runs
```{r}
summariseQ(tabulateQ(alist))
```
### Row labels
```{r}
popmap_rel <- popmap_relabeled[ popmap_relabeled$id %in% colnames(vcfR_panel_filt_rel@gt), ]

sites <- data.frame("id" = locations$UI.Barcode, "site" = locations$Region)
sites$site[is.na(sites$site)] <- "Unknown"

popmap_rel <- left_join(popmap_rel, sites)

rownames(alist[[1]]) <- popmap_rel$id

labels <-  data.frame(pop = as.character(popmap_rel$pop))
```
## View barplots K = 2 through 6
```{r}
# Organize
qlist <- alist[c(11,16,21,26,31)]

rownames(qlist$related_removed.2.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.2.1.Q))
rownames(qlist$related_removed.3.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.3.1.Q))
rownames(qlist$related_removed.4.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.4.1.Q))
rownames(qlist$related_removed.5.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.5.1.Q))
rownames(qlist$related_removed.6.1.Q) <- paste(labels$pop, rownames(qlist$related_removed.6.1.Q))

qlist$related_removed.4.1.Q <- data.frame("Cluster1" = qlist$related_removed.4.1.Q$Cluster3, "Cluster2" = qlist$related_removed.4.1.Q$Cluster2, "Cluster3" = qlist$related_removed.4.1.Q$Cluster4, "Cluster4" = qlist$related_removed.4.1.Q$Cluster1)

qlist$related_removed.5.1.Q <- data.frame("Cluster1" = qlist$related_removed.5.1.Q$Cluster1, "Cluster2" = qlist$related_removed.5.1.Q$Cluster4, "Cluster3" = qlist$related_removed.5.1.Q$Cluster5, "Cluster4" = qlist$related_removed.5.1.Q$Cluster2, "Cluster5" = qlist$related_removed.5.1.Q$Cluster3)

arrange(qlist$related_removed.6.1.Q, desc(Cluster6))

qlist$related_removed.6.1.Q <- data.frame("Cluster1" = qlist$related_removed.6.1.Q$Cluster6, "Cluster2" = qlist$related_removed.6.1.Q$Cluster3, "Cluster3" = qlist$related_removed.6.1.Q$Cluster4, "Cluster4" = qlist$related_removed.6.1.Q$Cluster2, "Cluster5" = qlist$related_removed.6.1.Q$Cluster1, "Cluster6" = qlist$related_removed.6.1.Q$Cluster5)
```
### Plot
```{r}
# Define your custom color palette
#"#fce5c9","#b5dee4","#3a7380","#f6b26b","#b45f06","#682502","#431801"

my_colors <- c("#6aa8b1","#f9d5a9","#431801","#eca255","#b45f06","#783513")  # Replace/add colors as needed for your K

# Plot with custom colors
p <- plotQ(
  qlist = qlist,
  imgoutput = "join",
  returnplot = TRUE,
  exportplot = FALSE,
  basesize = 10,
  grplab = labels,
  splabcol = "black",
  ordergrp = TRUE,
  subsetgrp = c("Kodiak",
  "SCAK",
  "NSEAK",
  "Unknown",
  "CSEAK",
  "SSEAK",
  "P. of Wales Is.",
  "Haida Gwaii"),
  grplabangle = 20,
  grplabsize = 5.5,
  grplabcol = "black",
  linesize = 0,
  pointsize = 0,
  grplabpos = 0.9,
  grplabspacer = 0,
  grplabjust = 1,
  grplabheight = 15,
  splabsize = 28,
  divgrp = "pop",
  divsize = 1,
  divtype = 1,
  clustercol = my_colors   # <--- custom color palette here
)

#c("Kodiak", "SCAK", "NSEAK", "Unknown", "CSEAK", "SSEAK", "B.C.", "Haida Gwaii")
png("figures/tissues_manuscript/SBTD_admixture.png", width = 6000, height = 6000, res = 600)
grid.arrange(p$plot[[1]])
dev.off()
```
# Locus variability and utility by population
## Format data
```{r}
popmap_var <- data.frame(id = popmap_relabeled$id, pop = as.character(popmap_relabeled$pop))
## Combine BC with SE since low sample size and groups as a population, put NA samples in the  SE/BC group
popmap_var$pop[popmap_var$pop == "Unknown"] <- "NSEAK"

popmap_var <- left_join(popmap_var, locations, by = c("id"="UI.Barcode")) %>% select(id, pop, Region)
popmap_var$Region[is.na(popmap_var$Region)] <- "NSEAK"

genind <- vcfR2genind(vcfR_panel_filt)
pop(genind) <- popmap_var$pop # Pop by Island/site or Region
genind_seppop <- seppop(genind)
```
## Locus stats by region
```{r}
pop_summary <- map_dfr(
  names(genind_seppop),
  function(pop_name) {
    g <- genind_seppop[[pop_name]]
    summ <- summary(g)
    maf <- minorAllele(g)
    
    tibble(
      Broad_region = pop_name,
      Locus = locNames(g),
      n = summ$n,
      Ho = summ$Hobs,
      MAF = as.numeric(maf),
      label = paste(pop_name," n=", summ$n, sep = "")
    )
  }
)

head(pop_summary)

pop_summary$Broad_region <- factor(pop_summary$Broad_region, levels = c("Kodiak", "SCAK", "NSEAK", "CSEAK", "SSEAK", "P. of Wales Is.", "Haida Gwaii"))
pop_summary$label <- factor(pop_summary$label, levels = unique(pop_summary$label[order(pop_summary$Broad_region)]))

pop_summary %>% 
  group_by(label) %>% 
  summarise(invar = sum(MAF <= 0),
            mean_MAF = mean(MAF) %>% round(digits = 2),
            sd_MAF = sd(MAF) %>% round(digits = 2),
            mean_Ho = mean(Ho) %>% round(digits = 2),
            sd_Ho = sd(Ho) %>% round(digits = 2))

popmap_var$pop <- factor(popmap_var$pop, levels = c("Kodiak", "SCAK", "NSEAK", "CSEAK", "SSEAK", "P. of Wales Is.", "Haida Gwaii"))

# Sample size by broad region/pop
popmap_var %>% 
  group_by(Region, pop) %>% 
  summarise(n = n()) %>% arrange(pop)
```
### Boxplot of observed het and maf for panel loci by region
```{r}
color_pal <- c("#f9d5a9", "#9ac9d0", "#3a7380", "#eca255", "#b45f06", "#783513", "#431801")

pop_summary$Broad_region <- factor(pop_summary$Broad_region, levels = c("Kodiak", "SCAK", "NSEAK", "CSEAK", "SSEAK", "P. of Wales Is.", "Haida Gwaii"))
pop_summary$label <- factor(pop_summary$label, levels = unique(pop_summary$label[order(pop_summary$Broad_region)]))

library(scales)  # for label_number()

# p1: Observed Heterozygosity
p1 <- ggplot(pop_summary, aes(x = label, y = Ho, fill = Broad_region)) +
  geom_boxplot(alpha = 0.9, color = "black") +  # 90% opacity, black outline
  scale_fill_manual(values = color_pal, guide = "none") +  # apply palette to fill
  scale_x_discrete(
  labels = function(x) sub("\\s+n=", "\nn=", x)
) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = NULL,
    y = "Observed Heterozygosity",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    axis.text.x = element_blank(),
    axis.title.y = element_text(size = 12)
  )

# p2: Minor Allele Frequency

p2 <- ggplot(pop_summary, aes(x = label, y = MAF, fill = Broad_region)) +
  geom_boxplot(alpha = 0.9, color = "black") +  # 90% opacity, black outline
  scale_fill_manual(values = color_pal, guide = "none") +  # apply palette to fill
  scale_x_discrete(
  labels = function(x) sub("\\s+n=", "\nn=", x)
) +
  scale_y_continuous(labels = label_number(accuracy = 0.01)) +
  labs(
    x = "Region",
    y = "Minor Allele Frequency"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 10),
    axis.title.y = element_text(size = 12)
  )
```
# Save boxplot
```{r}
png("figures/tissues_manuscript/SBTD_panel_het_maf_region_boxplots.png", width = 5000, height = 3000, res = 600)

grid.arrange(p1, p2, ncol = 1, heights = c(1,1.6))

dev.off()
```
### Number variable by region
```{r}
invar_summary_df <- pop_summary %>%
  filter(MAF == 0) %>% # Invariable loci
  group_by(Broad_region) %>%       # group by region
  summarise(n_invar_loci = n()) %>%      # count number of loci
  arrange(desc(n_invar_loci))           # sort by count

invar_summary_df

var_summary_df <- pop_summary %>%
  filter(Ho > 0.2) %>%
  filter(MAF > 0.2) %>% # Invariable loci
  group_by(Broad_region) %>%       # group by region
  summarise(n_loci = n()) %>%      # count number of loci
  arrange(n_loci)           # sort by count

var_summary_df

left_join(var_summary_df, invar_summary_df)
```
# PID/PIDSib
## Calculate PID and PIDsib by population/region
Original simulation script from Dana Morin, modified for multiple regions/populations
```{r}
library(dplyr)
library(tidyr)
library(purrr)

# Parameters
N <- 1000 # number of simulations
x <- 40   # max number of loci to simulate through

# Function to calculate PID and PIDsib for one population
calc_pid <- function(genind_obj, popname, N, x) {
  
  # allele frequencies
  freq <- makefreq(genind_obj) %>% as.data.frame()
  df <- colMeans(freq, na.rm = TRUE) %>% as.data.frame()
  row_names <- rownames(df)
  
  locus_names <- sub("^(([^:]+):[^:]+):.*$", "\\1", row_names)
  
  df$Marker <- locus_names
  df$allele <- ifelse(grepl("\\.0$", row_names), "X1", "X2")
  
  allele_freqs <- pivot_wider(df, 
                              id_cols = Marker,
                              names_from = allele, 
                              values_from = .)
  
  # format allele frequencies into list
  freqs <- allele_freqs
  al.freq.ls <- NULL
  for (i in 1:length(freqs$Marker)) {
    addon <- unname(unlist(c(freqs[i, -1])))
    addon <- addon[!is.na(addon)]
    addon <- list(addon)
    al.freq.ls <- c(al.freq.ls, addon)
  }
  
  # simulation
  PID.pop.tbl <- matrix(NA, nrow = x, ncol = 2)
  PID.pop.sib.tbl <- matrix(NA, nrow = x, ncol = 2)
  
  for (j in 1:x) { 
    PIpop.loci <- NULL
    PIpop.sib.loci <- NULL
    
    for (h in 1:N) {
      samp <- sample(al.freq.ls, j, replace = FALSE)
      PIpop <- NULL
      PIpop.sib <- NULL
      
      for (i in 1:length(samp)) {
        r1 <- sum(samp[[i]]^4)
        r2 <- outer(X = samp[[i]], Y = samp[[i]], FUN = function(X,Y)(2*X*Y)^2)
        r <- sum(r2[lower.tri(r2)]) + r1
        s <- 0.25 + (0.5 * sum(samp[[i]]^2)) + (0.5 * sum(samp[[i]]^2)^2) -
          (0.25 * sum(samp[[i]]^4))
        
        PIpop <- c(PIpop, r)
        PIpop.sib <- c(PIpop.sib, s)
      }
      PIpop.loci <- c(PIpop.loci, prod(PIpop))
      PIpop.sib.loci <- c(PIpop.sib.loci, prod(PIpop.sib))
    }
    
    PID.pop.tbl[j, 1] <- j
    PID.pop.tbl[j, 2] <- mean(PIpop.loci)
    PID.pop.sib.tbl[j, 1] <- j
    PID.pop.sib.tbl[j, 2] <- mean(PIpop.sib.loci)
  }
  
  PID.pop.tbl <- as.data.frame(PID.pop.tbl)
  colnames(PID.pop.tbl) <- c("Loci", "Avg.PID")
  
  PID.pop.sib.tbl <- as.data.frame(PID.pop.sib.tbl)
  colnames(PID.pop.sib.tbl) <- c("Loci", "Avg.PIDsib")
  
  # merge PID and PIDsib, add population column
  all_PID <- left_join(PID.pop.sib.tbl, PID.pop.tbl, by = "Loci") %>%
    mutate(Population = popname)
  
  return(all_PID)
}

# ---- Run across all populations ----
all_PID <- map2_dfr(genind_seppop, names(genind_seppop),
                    ~calc_pid(.x, .y, N, x))

# Final output
library(dplyr)

all_PID$Population <- factor(all_PID$Population, levels = rev(c("Haida Gwaii", "P. of Wales Is.","SSEAK", "CSEAK", "NSEAK", "SCAK", "Kodiak"))) 

all_PID %>%
  group_by(Population) %>%
  filter(Avg.PIDsib < 0.001) %>%
  slice(1) %>%
  select(Population, Loci) %>%
  mutate(label = paste0(Population, " = ", Loci)) %>%
  pull(label) %>%
  paste(collapse = ", ")
```
# Plot PID and PIDsib
```{r}
pid_color_pal <- rev(c("#431801", "#783513","#b45f06","#eca255","#3a7380","#9ac9d0","#f9d5a9"))

all_PID$Population <- factor(all_PID$Population, levels = rev(c("Haida Gwaii", "P. of Wales Is.","SSEAK", "CSEAK", "NSEAK", "SCAK", "Kodiak"))) 


p_PID <- ggplot(all_PID, aes(x = Loci, y = Avg.PID, color = Population)) +
  geom_line(size = 1) +
  geom_hline(aes(yintercept = 0.001, linetype = "PID = 0.001"), color = "black") +
  scale_linetype_manual(
    values = c("PID = 0.001" = "dashed")
  ) +
  scale_x_continuous(limits = c(0, 20)
  ) +
  coord_cartesian(ylim = c(0, 0.1)
  ) +
  scale_color_manual(values = pid_color_pal) +
  labs(
  #  title = "PID with Panel, Nsim = 1000",
    x = "Number of Loci",
    y = "Probability of Identity",
    color = NULL,
    linetype = NULL
  ) +
  theme_minimal()

p_PIDsib <- ggplot(all_PID, aes(x = Loci, y = Avg.PIDsib, color = Population)) +
  geom_line(size = 1.5) +
  geom_hline(
    aes(yintercept = 0.001, linetype = "pid_sibs"),
    color = "black"
  ) +
  scale_linetype_manual(
    values = c(pid_sibs = "dashed"),
    labels = c(expression(PID[SIBS] == 0.001))
  ) +
  scale_x_continuous(limits = c(0, 40)) +
  coord_cartesian(ylim = c(0, 0.1)) +
  scale_color_manual(values = pid_color_pal) +
  labs(
    x = "Number of Loci",
    y = "Probability of Identity",
    color = NULL,
    linetype = NULL
  ) +
  theme_minimal(base_size = 14)
```
### Save plots
```{r}
png("figures/tissues_manuscript/SBTD_PIDsib.png", width = 5000, height = 3000, res = 600)

p_PIDsib

dev.off()

png("figures/tissues_manuscript/SBTD_PID.png", width = 5000, height = 3000, res = 600)

p_PID

dev.off()
```
# List of variable loci by region
```{r}
# Assess if any loci are out of HWE
HWE_SC <- data.frame(hw.test(genind_seppop$SCAK)) 
HWE_nSE <- data.frame(hw.test(genind_seppop$`NSEAK`))
HWE_HG <- data.frame(hw.test(genind_seppop$`Haida Gwaii`))
HWE_KA <- data.frame(hw.test(genind_seppop$Kodiak))
HWE_POW <- data.frame(hw.test(genind_seppop$`P. of Wales Is.`))
HWE_mSE <- data.frame(hw.test(genind_seppop$`CSEAK`))
HWE_sSE <- data.frame(hw.test(genind_seppop$`CSEAK`))

# P adjust with Benjamini Hochberg FDR method
HWE_SC_subset <- HWE_SC[p.adjust(HWE_SC$Pr.exact, method = "BH") < 0.05, ] # none out of HWE
HWE_nSE_subset <- HWE_nSE[p.adjust(HWE_nSE$Pr.exact, method = "BH") < 0.05, ]
HWE_HG_subset <- HWE_HG[p.adjust(HWE_HG$Pr.exact, method = "BH") < 0.05, ] # none out of HWE
HWE_KA_subset <- HWE_KA[p.adjust(HWE_KA$Pr.exact, method = "BH") < 0.05, ]
HWE_POW_subset <- HWE_POW[p.adjust(HWE_POW$Pr.exact, method = "BH") < 0.05, ]  # none out of HWE
HWE_mSE_subset <- HWE_mSE[p.adjust(HWE_mSE$Pr.exact, method = "BH") < 0.05, ]
HWE_sSE_subset <- HWE_sSE[p.adjust(HWE_sSE$Pr.exact, method = "BH") < 0.05, ]

HWE_nSE_subset$Region <- "nSE"
HWE_KA_subset$Region <- "KA"
HWE_mSE_subset$Region <- "mSE"
HWE_sSE_subset$Region <- "sSE"

HWE <- data.frame(Locus = c(rownames(HWE_nSE_subset), rownames(HWE_KA_subset), rownames(HWE_mSE_subset), rownames(HWE_sSE_subset)), info = c(HWE_nSE_subset$Region, HWE_KA_subset$Region, HWE_mSE_subset$Region, HWE_sSE_subset$Region))

High_het <- pop_summary %>% filter(Ho > 0.7) %>% select(Locus) %>% unique()
High_het$info <- "High_het"

HWE_high_het <- rbind(HWE, High_het) %>% arrange(Locus)

unique(HWE_high_het$Locus) # remove these from recommended loci

var_loci_df <- pop_summary %>%
  filter(Ho > 0.2) %>%
  filter(MAF > 0.2) %>% # Invariable loci
  group_by(Broad_region)      # group by region

recommended <- var_loci_df %>% filter(!Locus %in% unique(HWE_high_het$Locus))
```
# Create recommended loci table
```{r}
F_R_seq <- read.csv("data/F_R_sequences.csv")
F_R_seq$CHROM_POS <- paste(F_R_seq$CHROM,F_R_seq$POS, sep = "_")

locus_info <- data.frame(vcfR_panel_filt@fix) %>% select(CHROM:ALT)

rec_locus_info <- locus_info[locus_info$ID %in% unique(recommended$Locus), ]
rec_locus_info$CHROM_POS <- paste(rec_locus_info$CHROM, rec_locus_info$POS, sep = "_")

rec_F_R <- left_join(rec_locus_info, F_R_seq, by = "CHROM_POS")
recommended_table <- left_join(recommended, rec_F_R, by = c("Locus"="ID")) %>% select(Broad_region:MAF,CHROM.x:ALT, Forward.sequence, Reverse.sequence)

colnames(recommended_table) <- c("Broad Region", "Locus ID", "N Samples", "Observed Heterozygosity", "Minor Allele Frequency", "Chromosome", "Position", "Reference Allele", "Alternate Allele", "Forward Sequence", "Reverse Sequence")

write.csv(recommended_table, file = "output/recommended_locus_set_by_region.csv")
```

# Fecal Opt matching test
```{r}
vcfR_f_opt <- read.vcfR("../Fecal_samples/Fecal_Opt_1/Populations_Runs/Populations_EDeer_R50_mac3/populations.snps.vcf")

panel_snps <- read.csv("/mnt/ceph/estacy/SBTDeer/R_analysis/data/Eacker_Deer_248_Loci_Names.csv", header = F)

# Whitelist snps panel was designed for only
panel_snps <- panel_snps %>%
  separate(V1, into = c("CHROM", "POS"), sep = "_")

# Create combined keys
vcf_keys <- paste(vcfR_f_opt@fix[, "CHROM"], vcfR_f_opt@fix[, "POS"], sep = "_")
panel_keys <- paste(panel_snps$CHROM, panel_snps$POS, sep = "_")

# Filter based on matching keys
vcfR_f_opt_panel <- vcfR_f_opt[vcf_keys %in% panel_keys, ]
```
```{r}
sample_data <- read.csv("data/SBTD-ED Metadata 2018-2025.csv")
vcfR_f_opt_panel

vcfR_f_opt_dp <- hard_filter(vcfR_f_opt_panel, depth = 5, gq = 30)

vcfR_f_opt_miss <- missing_by_sample(vcfR_f_opt_dp, cutoff = 0.8)

vcfR_f_opt_miss_snp <- missing_by_snp(vcfR_f_opt_miss, cutoff = 0.1)

fecal_opt_samples <- data.frame(Lab.ID = sub(".*_","", colnames(vcfR_f_opt_miss@gt[,2:85])) %>% unique())

left_join(fecal_opt_samples, sample_data)

data.frame(t(extract.gt(vcfR_f_opt_miss, element = "GT")))
```
```{r}
popmap_fecal <- data.frame(id = colnames(vcfR_f_opt_miss_snp@gt[,2:85]))

# Format paired genotype data frame
paired_gt_df_fecal <- format_paired_gt_df(vcfR_f_opt_miss_snp, popmap_fecal)
    
# Calculate error rate
error_rate_fecal <- calculate_gt_pairs_errors(paired_gt_df_fecal)

hist(error_rate_fecal$summary_by_pair$Mismatch_Rate, breaks = 10, main = "Fecal DNA Replicate Mismatch Rate", xlab = "Mismatch Rate")

mean(error_rate_fecal$summary_by_pair$Mismatch_Rate)
```
## Create consensus genotyes
```{r}
fecal_popmap <- data.frame(id = sub(
  "Gtseq_\\d+_([A-Z]\\d+)_SBTD_((?:FOPT|FP)\\d+)_([A-Za-z0-9]+)",
  "\\3_\\2_\\1",
  popmap_fecal$id,
  ignore.case = TRUE,
  perl = TRUE
))

colnames(vcfR_f_opt_miss_snp@gt) <- c("FORMAT", fecal_popmap$id) # Re-label vcfR

vcfR <- vcfR_f_opt_miss_snp

# Extract genotypes
rep_GTs <- data.frame(locus_id = rownames(extract.gt(vcfR, element = "GT")),
                      extract.gt(vcfR, element = "GT"),
                      row.names = NULL, check.names = F)

# Extract depths
rep_DPs <- data.frame(locus_id = rownames(extract.gt(vcfR, element = "DP")),
                      extract.gt(vcfR, element = "DP"),
                      row.names = NULL, check.names = F)
rep_DPs[2:ncol(rep_DPs)] <- lapply(rep_DPs[2:ncol(rep_DPs)], as.numeric)

# Transpose: rows = samples, columns = loci, first column = sample IDs
transpose_gt <- function(df) {
  locus_ids <- df[[1]]
  mat <- t(df[, -1, drop = FALSE])
  result <- data.frame(sample_id = rownames(mat), mat, row.names = NULL, check.names = FALSE)
  colnames(result)[-1] <- locus_ids
  result
}

rep_GTs_t <- transpose_gt(rep_GTs)
rep_DPs_t <- transpose_gt(rep_DPs)

results <- consensus_genotypes(rep_GTs_t, rep_DPs_t)
consensus_gt_df    <- results$genotypes
consensus_depth_df <- results$depths
```
# Run matching method
### Format for matching code
```{r}
geno_matrix <- convert_gt_to_numeric(consensus_gt_df)

consensus_gt_df %>% 
  group_by(sample_id) %>%
  summarise(na_count = rowSums(across(everything(), ~ is.na(.))),
            prop_miss = na_count/217) %>% arrange(desc(na_count))

pairs <- c("KYZXG", "KYZY5")
count_allele_mismatches(pairs, geno_matrix)
```
### Function to compute allele frequencies and compute
```{r}
compute_allele_freq <- function(geno_matrix) {
  
  L <- ncol(geno_matrix) # number of loci
  p <- numeric(L)
  
  for (j in seq_len(L)) {
    
    g <- geno_matrix[, j]
    g <- g[!is.na(g)]
    
    if (length(g) == 0) {
      p[j] <- NA
      next
    }
    
    nAA <- sum(g == 1L)
    nAB <- sum(g == 2L)
    n   <- length(g)
    
    p[j] <- (2*nAA + nAB) / (2*n)
  }
  
  return(p)
}

allele_freq <- compute_allele_freq(geno_matrix)
```
## Locus error rates for matching calculations
```{r}
ERR.locus <- round(error_rate_fecal$summary_by_locus$Both_Mismatch_Rate, digits = 4) # Individually calculated based on replicates, based on same depth and missingness filtering

min_error <- 1e-4  # Add in a minimum error to allow for some error in 0 error calc loci
ERR.locus <- pmax(ERR.locus, min_error)

gamma <- 1 - sqrt(1 - ERR.locus)
```
## Functions to calculate probabilities of observing a pair of latent diploid genotypes and per locus joint probability array
```{r}
# Two homozygous genotypes, two shared alleles 
# Arguments: a single allele frequency corresponds to the matching allele 
        hom.hom2.SI <- function(f){ 
          return(f[1]^2) 
          } # End joint genotype probability calculation for observing ii/ii given same individual 
        hom.hom2.U <- function(f){ 
          return(f[1]^4) 
          } # End joint genotype probability calculation for observing ii/ii given two unrelated individuals 
        hom.hom2.PO <- function(f){ 
          return(f[1]^3) 
          } # End joint genotype probability calculation for observing ii/ii given a parent offspring relationship 
        hom.hom2.FS <- function(f){ 
          return(0.25*f[1]^2 + 0.5*f[1]^3 + 0.25 * f[1]^4) 
          } # End joint genotype probability calculation for observing ii/ii given two full siblings 
         
# Two homozygous genotypes, zero shared alleles 
# Arguments: vector of two allele frequencies which correspond to one homozygote and second to the other, order doesn't matter 
        hom.hom0.SI <- function(f){ 
          return(0) 
          } # End joint genotype probability calculation for observing ii/jj given same individual 
        hom.hom0.U <- function(f){ 
          return(f[1]^2 * f[2]^2) 
          } # End joint genotype probability calculation for observing ii/jj given two unrelated individuals 
        hom.hom0.PO <- function(f){ 
          return(0) 
          } # End joint genotype probability calculation for observing ii/jj given a parent offspring relationship 
        hom.hom0.FS <- function(f){ 
          return(0.25*f[1]^2 * f[2]^2) 
          } # End joint genotype probability calculation for observing ii/jj given two full siblings 
           
      # One homozygote, one heterozygote, one shared allele  
      # Arguments: vector of two allele frequencies, here the first allele frequency must correspond to the  
      # homozygous genotype and the second to the novel allele in the heterzygote, e.g. ii/ij => f = c(f_i,f_j). 
        hom.het1.SI <- function(f){ 
          return(0) 
          } # End joint genotype probability calculation for observing ii/ij given same individual 
        hom.het1.U <- function(f){ 
          return(2*f[1]^3 * f[2]) 
          } # End joint genotype probability calculation for observing ii/ij given two unrelated individuals 
        hom.het1.PO <- function(f){ 
          return(f[1]^2 * f[2]) 
          } # End joint genotype probability calculation for observing ii/ij given a parent offspring relationship 
        hom.het1.FS <- function(f){ 
          return(0.5*f[1]^2 * f[2] + 2*0.25*f[1]^3 * f[2]) 
          } # End joint genotype probability calculation for observing ii/ij given two full siblings 
 
# Two heterozygote genotypes, two shared alleles 
# Arguments: a vector of two allele frequencies, the first allele corresponds to one allele in the heterozygote, and the second the other, order doesn't matter 
        het.het2.SI <- function(f){ 
          return(2*f[1] * f[2]) 
          } # End joint genotype probability calculation for observing ij/ij given same individual 
        het.het2.U <- function(f){ 
          return(4*f[1]^2 * f[2]^2) 
          } # End joint genotype probability calculation for observing ij/ij given two unrelated individuals 
        het.het2.PO <- function(f){ 
          return(f[1] * f[2] * (f[1] + f[2]) ) 
          } # End joint genotype probability calculation for observing ij/ij given a parent offspring relationship 
        het.het2.FS <- function(f){ 
          return( (2*0.25*f[1] * f[2]) + (0.5*f[1] * f[2] * (f[1] + f[2]) ) + (4*0.25*f[1]^2 * f[2]^2)) 
          } # End joint genotype probability calculation for observing ij/ij given two full siblings 

build_joint_snp <- function(p) {
  
  # allele frequencies
  f <- c(a = p, b = 1 - p)
  
  # 3 genotypes × 3 genotypes × 4 relationships (SI: same indv, U: unrelated, PO: parent-offspring, FS: full sibs)
  J <- array(0,
             dim = c(3, 3, 4),
             dimnames = list(
               g1 = c("aa","ab","bb"),
               g2 = c("aa","ab","bb"),
               R  = c("SI","U","PO","FS")
             ))
  
  ### 1) aa / aa
  J[1,1,] <- c(
    hom.hom2.SI(f["a"]),
    hom.hom2.U(f["a"]),
    hom.hom2.PO(f["a"]),
    hom.hom2.FS(f["a"])
  )
  
  ### 2) bb / bb
  J[3,3,] <- c(
    hom.hom2.SI(f["b"]),
    hom.hom2.U(f["b"]),
    hom.hom2.PO(f["b"]),
    hom.hom2.FS(f["b"])
  )
  
  ### 3) aa / bb  and  bb / aa
  val <- c(
    hom.hom0.SI(f),
    hom.hom0.U(f),
    hom.hom0.PO(f),
    hom.hom0.FS(f)
  )
  J[1,3,] <- val
  J[3,1,] <- val
  
  ### 4) aa / ab  and  ab / aa
  val <- c(
    hom.het1.SI(c(f["a"], f["b"])),
    hom.het1.U(c(f["a"], f["b"])),
    hom.het1.PO(c(f["a"], f["b"])),
    hom.het1.FS(c(f["a"], f["b"]))
  )
  J[1,2,] <- val
  J[2,1,] <- val
  
  ### 5) bb / ab  and  ab / bb
  val <- c(
    hom.het1.SI(c(f["b"], f["a"])),
    hom.het1.U(c(f["b"], f["a"])),
    hom.het1.PO(c(f["b"], f["a"])),
    hom.het1.FS(c(f["b"], f["a"]))
  )
  J[3,2,] <- val
  J[2,3,] <- val
  
  ### 6) ab / ab
  J[2,2,] <- c(
    het.het2.SI(f),
    het.het2.U(f),
    het.het2.PO(f),
    het.het2.FS(f)
  )
  
  return(J)
}
```
## Build joint array for all loci based on calculated allele frequencies
```{r}
L <- length(allele_freq)

joint_array <- array(NA,
                     dim = c(L, 3, 3, 4),
                     dimnames = list(
                       locus = NULL,
                       g1 = c("aa","ab","bb"),
                       g2 = c("aa","ab","bb"),
                       R  = c("SI","U","PO","FS")
                     ))

for (j in seq_len(L)) {
  
  p <- allele_freq[j]
  
  if (is.na(p)) next
  
  joint_array[j,,,] <- build_joint_snp(p)
}
```
## Per pair likelihood calculation
```{r}
# Mismatch matrix
mismatch <- matrix(
  c(0,1,2,
    1,0,1,
    2,1,0),
  nrow = 3,
  byrow = TRUE
)

# Error log probability
logP_error <- function(gamma, d, n_a) {
  d * log(gamma) + (n_a - d) * log(1 - gamma)
}

# Per pair likelihood function
compute_pair_logLR <- function(i, k,
                               geno_matrix,
                               joint_array,
                               gamma,
                               mismatch) {
  
  L <- ncol(geno_matrix)
  logL <- rep(0, 4)  # SI, U, PO, FS
  
  for (j in seq_len(L)) {
    
    g1 <- geno_matrix[i, j]
    g2 <- geno_matrix[k, j]
    
    if (is.na(g1) || is.na(g2)) next
    
    gamma_j <- gamma[j]
    
    # Number of alleles observed = 2 per individual
    n_a <- 2
    
    # True genotype combinations (1:3 × 1:3)
    J <- joint_array[j,,,]  # 3×3×4
    
    # Accumulator per relationship
    Lj <- rep(0, 4)
    
    # Sum over k1,k2 (true genotypes)
    for (k1 in 1:3) {
      for (k2 in 1:3) {
        
        d1 <- mismatch[g1, k1]
        d2 <- mismatch[g2, k2]
        
        logPe1 <- logP_error(gamma_j, d1, n_a)
        logPe2 <- logP_error(gamma_j, d2, n_a)
        
        # joint true genotype probability
        logJoint <- log(J[k1, k2, ])
        
        # accumulate in probability scale
        Lj <- Lj + exp(logJoint + logPe1 + logPe2)
      }
    }
    
    logL <- logL + log(Lj)
  }
  
  # log likelihood ratio
  logLR <- logL[1] - max(logL[2:4])
  
  return(logLR)
}
```
## Compute all pairwise LogLR
### Step 1: Vectorized pair function
```{r}
compute_pair_logLR_vec <- function(i, k,
                                   geno_matrix,
                                   joint_array,
                                   gamma,
                                   mismatch) {
  
  g1 <- geno_matrix[i, ]
  g2 <- geno_matrix[k, ]
  
  # Only compute loci where both are genotyped
  both <- !is.na(g1) & !is.na(g2)
  if (sum(both) == 0) return(-Inf)
  
  g1 <- g1[both]; g2 <- g2[both]
  gamma_j <- gamma[both]
  J <- joint_array[both, , , , drop = FALSE]  # n_loci × 3 × 3 × 4
  
  n_a <- 2
  
  # Precompute error log-probs for all loci and all true genotype combos
  # d1[j, k1] = mismatch between observed g1[j] and true genotype k1
  d1 <- mismatch[g1, ]   # n_loci × 3
  d2 <- mismatch[g2, ]   # n_loci × 3
  
  # log P(error) for each locus × true genotype: n_loci × 3
  logPe1 <- d1 * log(gamma_j) + (n_a - d1) * log(1 - gamma_j)  
  logPe2 <- d2 * log(gamma_j) + (n_a - d2) * log(1 - gamma_j)  
  
  # Sum over all 9 true genotype combinations (k1, k2)
  # Result: n_loci × 4
  Lj <- matrix(0, nrow = sum(both), ncol = 4)
  
  for (k1 in 1:3) {
    for (k2 in 1:3) {
      # J[, k1, k2, ] is n_loci × 4
      logJoint <- log(J[, k1, k2, ])           # n_loci × 4
      pe       <- logPe1[, k1] + logPe2[, k2]  # n_loci (scalar per locus)
      Lj <- Lj + exp(logJoint + pe)             # broadcast pe across 4 cols
    }
  }
  
  # Sum log-likelihoods across loci
  logL <- colSums(log(Lj))  # length 4
  
  logLR <- logL[1] - max(logL[2:4])
  return(logLR)
}
```
## Step 2: Chunk pairs to reduce overhead and parallelize to run faster
```{r}
library(doParallel)
library(foreach)

# Set up cluster
n_cores <- parallel::detectCores() - 1
cl <- makeCluster(n_cores)
registerDoParallel(cl)

n <- nrow(geno_matrix)

# Generate all unique pairs
pairs <- do.call(rbind, lapply(1:(n-1), function(i) data.frame(i = i, k = (i+1):n)))

# Chunk pairs to reduce parallelization overhead
chunk_size <- 1000
pair_list <- split(1:nrow(pairs), ceiling(seq_len(nrow(pairs)) / chunk_size))

start_time <- Sys.time()

# Parallel loop over chunks — no .export needed, variables auto-exported
results_list <- foreach(
  chunk = pair_list,
  .combine = c  # combine as a flat list, not rbind
) %dopar% {
  lapply(chunk, function(idx) {
    row <- pairs[idx, ]
    val <- compute_pair_logLR_vec(row$i, row$k,
                                  geno_matrix,
                                  joint_array,
                                  gamma,
                                  mismatch)
    list(i = row$i, k = row$k, logLR = val)
  })
}

stopCluster(cl)

end_time <- Sys.time()
cat("Time elapsed:", difftime(end_time, start_time, units = "mins"), "mins\n")

# Extract results safely from list
i_idx   <- sapply(results_list, `[[`, "i")
k_idx   <- sapply(results_list, `[[`, "k")
logLR_vals <- sapply(results_list, `[[`, "logLR")

# Rebuild symmetric matrix
logLR_matrix <- matrix(-Inf, n, n)
logLR_matrix[cbind(i_idx, k_idx)] <- logLR_vals
logLR_matrix[cbind(k_idx, i_idx)] <- logLR_vals
diag(logLR_matrix) <- -Inf

rownames(logLR_matrix) <- rownames(geno_matrix)
colnames(logLR_matrix) <- rownames(geno_matrix)

adj <- logLR_matrix > 0
```
## Step 3: Cluster with igraph
```{r}
library(igraph)

g <- graph_from_adjacency_matrix(adj,
                                 mode = "undirected",
                                 diag = FALSE)

clusters <- components(g)
```
# Make sample name and matching data frame
```{r}
sample_names <- rownames(geno_matrix)

out_df <- data.frame(
  Sample = sample_names,
  Individual_ID = clusters$membership,
  stringsAsFactors = FALSE
)

# Individual IDs
out_df %>% arrange(Individual_ID)

unique(out_df$Individual_ID)
 
df <- left_join(out_df, sample_data, by = c("Sample"="Lab.ID")) %>% select(Sample, Individual_ID, Field.ID, Date, Site, TransectID, qPCR.Mean..ng.uL.) %>% arrange(Individual_ID)

write.csv(df, "output/fecal_opt_matching.csv")

freq_matches <- data.frame(table(out_df$Individual_ID)) %>% arrange(desc(Freq))

freq_matches$Var1 <- as.numeric(freq_matches$Var1) 

ggplot(freq_matches, aes(x=Freq)) +
  geom_histogram(bins = 3, colour = "black") +
  theme(text=element_text(size=15)) +
  xlab("Number of Sample Captures") +
  ylab("Number of Individuals")
```
